---
title: "Basic Statistics and Maps"
format: html
toc: true
---

## Summary

The goal of this page is to provide the map and a basic understanding of the data I have. This is to ensure visuals for publication have the same themeing but also to ensure that my data management aligns well with expectations and that there are no errors. If you see errors, please let me know!

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("basic_stats")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false
#| results: hide

# Load and process data using the common function
data <- load_ose_data(cache_processed = TRUE)

# Save basic dataset summary
save_data_table(
  data |> 
    summarise(
      n_farmers = n_distinct(farmer),
      n_regions = n_distinct(region),
      n_missions = n_distinct(mission_number),
      total_observations = n(),
      mean_ose_count = mean(ose_count, na.rm = TRUE),
      mean_temperature = mean(temperature, na.rm = TRUE),
      mean_ground_cover = mean(percent_ground_cover, na.rm = TRUE),
      mean_yield = mean(rendement_en_kg_ha, na.rm = TRUE)
    ),
  "dataset_summary"
)

# Basic statistics around dataset

```{r basic-statistics}
#| echo: false
#| message: false
#| warning: false

summary_table <- table_variable_summary(data)

# Also save the detailed variable summary
if (exists("summary_table")) {
  # Extract the underlying data if it's a gt table
  tryCatch({
    if ("gt_tbl" %in% class(summary_table)) {
      # For gt tables, we need to recreate the summary data
      summary_data <- data |>
        select(where(is.numeric)) |>
        summarise(across(everything(), list(
          min = ~min(.x, na.rm = TRUE),
          max = ~max(.x, na.rm = TRUE), 
          mean = ~mean(.x, na.rm = TRUE),
          median = ~median(.x, na.rm = TRUE),
          sd = ~sd(.x, na.rm = TRUE)
        ))) |>
        pivot_longer(everything(), names_to = "variable_stat", values_to = "value") |>
        separate(variable_stat, into = c("variable", "statistic"), sep = "_(?=[^_]*$)")
      
      save_data_table(summary_data, "variable_summary_statistics")
    }
  }, error = function(e) {
    message("Could not save summary table data: ", e$message)
  })
}

# Display the summary table
summary_table
```

# Map of Senegal and regions

```{r senegal-map}
#| echo: false
#| message: false
#| warning: false

senegal_map <- plot_senegal_map(
  regions_of_interest = STUDY_REGIONS,
  show_legend = FALSE,
  roi_color = "tomato",
  other_color = "grey80",
  label_size = 4
)

# Display the map
senegal_map
```