---
title: "Basic Statistics and Maps"
format: html
toc: true
---

## Summary

The goal of this page is to provide the map and a basic understanding of the data I have. This is to ensure visuals for publication have the same themeing but also to ensure that my data management aligns well with expectations and that there are no errors. If you see errors, please let me know!

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("basic_stats")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false
#| results: hide

# Load and process data using the common function
data <- load_ose_data(cache_processed = TRUE)

# Save basic dataset summary
save_data_table(
  data |> 
    summarise(
      n_farmers = n_distinct(farmer),
      n_regions = n_distinct(region),
      n_missions = n_distinct(mission_number),
      total_observations = n(),
      mean_ose_count = mean(ose_count, na.rm = TRUE),
      mean_temperature = mean(temperature, na.rm = TRUE),
      mean_ground_cover = mean(percent_ground_cover, na.rm = TRUE),
      mean_yield = mean(rendement_en_kg_ha, na.rm = TRUE)
    ),
  "dataset_summary"
)

# Basic statistics around dataset

#| echo: false
#| message: false
#| warning: false

summary_table <- table_variable_summary(data)

# Also save the detailed variable summary
if (exists("summary_table")) {
  # Extract the underlying data if it's a gt table
  tryCatch({
    if ("gt_tbl" %in% class(summary_table)) {
      # For gt tables, we need to recreate the summary data
      summary_data <- data |>
        select(where(is.numeric)) |>
        summarise(across(everything(), list(
          min = ~min(.x, na.rm = TRUE),
          max = ~max(.x, na.rm = TRUE), 
          mean = ~mean(.x, na.rm = TRUE),
          median = ~median(.x, na.rm = TRUE),
          sd = ~sd(.x, na.rm = TRUE)
        ))) |>
        pivot_longer(everything(), names_to = "variable_stat", values_to = "value") |>
        separate(variable_stat, into = c("variable", "statistic"), sep = "_(?=[^_]*$)")
      
      save_data_table(summary_data, "variable_summary_statistics")
    }
  }, error = function(e) {
    message("Could not save summary table data: ", e$message)
  })
}

# Display the summary table
summary_table
```

# Map of Senegal and regions

```{r senegal-map}
#| echo: false
#| message: false
#| warning: false

senegal_map <- plot_senegal_map(
  regions_of_interest = STUDY_REGIONS,
  show_legend = FALSE,
  roi_color = "tomato",
  other_color = "grey80",
  label_size = 4
)

# Display the map
senegal_map
```


# Mean Annual precipitation 

Mean annual precipitation (MAP) for Senegal was calculated using the CHIRPS dataset, which provides high-resolution (approximately 5 km) gridded precipitation estimates. We extracted CHIRPS daily precipitation data for Senegal for the most recent 10-year period (2015–2024). The data were spatially clipped to the national boundary, summed annually, and then averaged across years to produce a single raster layer representing mean annual precipitation.


```{r}
library(terra)
library(tidyterra)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(dplyr)

# 1. Import the MAP TIFF
map_raster <- rast("data/processed/maps/senegal_map_2015_2024_chirps.tif")

# 2. Get Senegal boundary and regions
senegal <- ne_states(country = "Senegal", returnclass = "sf")

# 3. Select regions of interest
regions_of_interest <- c("Saint-Louis", "Thiès", "Fatick", "Kaffrine")
roi_sf <- senegal %>% filter(name %in% regions_of_interest)
other_sf <- senegal %>% filter(!name %in% regions_of_interest)

# 4. Crop and mask raster to Senegal
map_raster_senegal <- crop(map_raster, vect(senegal))
map_raster_senegal <- mask(map_raster_senegal, vect(senegal))

# 5. Mask raster to only regions of interest for plotting
roi_union <- st_union(roi_sf)
map_raster_roi <- mask(map_raster_senegal, vect(roi_union))

# 6. Plot
ggplot() +
  geom_sf(data=senegal) +
  geom_spatraster(data = map_raster_senegal, aes(fill = senegal_map_2015_2024_chirps), na.rm = TRUE) +
  geom_sf(data = senegal, fill = NA, color = "grey", size = 0.6) +
  scale_fill_viridis_c(name = "MAP (mm)", na.value = NA,option="C") +
  theme_minimal() +
  labs(title = "Mean Annual Precipitation (2015–2024)",
       subtitle = "Saint-Louis, Thiès, Fatick, and Kaffrine highlighted") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())


```


## Just the regions


```{r}

ggplot() +
  geom_sf(data = other_sf, fill = "grey90", color = "grey", size = 0.2) +
  geom_spatraster(data = map_raster_roi, aes(fill = senegal_map_2015_2024_chirps), na.rm = TRUE) +
  geom_sf(data = roi_sf, fill = NA, color = "grey", size = 0.6) +
  geom_sf_text(data = roi_sf, aes(label = name), size = 4, fontface = "bold", color = "black") +
  scale_fill_viridis_c(name = "MAP (mm)", na.value = NA, option = "C") +
  theme_minimal() +
  labs(title = "Mean Annual Precipitation (2015–2024)",
       subtitle = "Saint-Louis, Thiès, Fatick, and Kaffrine highlighted") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())

```



```{r}
library(scales) # for comma formatting
roi_sf$mean_map <- terra::extract(map_raster, vect(roi_sf), fun = mean, na.rm = TRUE)[,2]
# 1. Create label with region name and mean MAP value
roi_sf$label <- sprintf(
  "%s\n(%s mm)",
  roi_sf$name,
  comma(round(roi_sf$mean_map, 0))
)

# 2. Choose a discrete color palette with good contrast
region_palette <- c(
  "Saint-Louis" = "#1b9e77",
  "Thiès"       = "#d95f02",
  "Fatick"      = "#7570b3",
  "Kaffrine"    = "#e7298a"
)

# 3. Plot with labels and custom colors
ggplot() +
  geom_sf(data = other_sf, fill = "grey90", color = "grey", size = 0.2) +
  geom_sf(data = roi_sf, aes(fill = name), color = "black", size = 0.6) +
  geom_sf_text(
    data = roi_sf,
    aes(label = label),
    size = 4.5, fontface = "bold",
    color = "white", # white text for contrast
    bg.color = "black", # if using ggtext or geom_sf_label, otherwise omit
    bg.r = 0.15
  ) +
  scale_fill_manual(
    name = NULL,
    values = region_palette,
    guide = "none"
  ) +
  theme_minimal() +
  labs(
    title = "Mean Annual Precipitation by Region (2015–2024)",
    subtitle = "Saint-Louis, Thiès, Fatick, and Kaffrine"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

# 2021 rainfall by month

```{r setup-roi-union}

months <- c("july", "august", "september", "october")
raster_paths <- paste0("data/processed/maps/senegal_precipitation_2021_", months, "_chirps.tif")
regions_of_interest <- c("Saint-Louis", "Thiès", "Fatick", "Kaffrine")
roi_sf <- senegal %>% filter(name %in% regions_of_interest)
other_sf <- senegal %>% filter(!name %in% regions_of_interest)
roi_union <- st_union(roi_sf)
```


```{r}
library(terra)

# Calculate global min/max for all rasters
rainfall_values <- lapply(raster_paths, function(path) {
  r <- rast(path)
  values(r)[!is.na(values(r))]
})
all_values <- unlist(rainfall_values)
rainfall_min <- min(all_values)
rainfall_max <- max(all_values)

# Helper function for full Senegal plot
plot_month_full <- function(month, raster_path, senegal, min_val, max_val) {
  raster <- rast(raster_path)
  raster_senegal <- crop(raster, vect(senegal)) |> mask(vect(senegal))
  ggplot() +
    geom_sf(data = senegal) +
    geom_spatraster(data = raster_senegal, aes(fill = !!sym(paste0("senegal_precipitation_2021_", month, "_chirps"))), na.rm = TRUE) +
    geom_sf(data = senegal, fill = NA, color = "grey", size = 0.6) +
    scale_fill_viridis_c(
      name = "Rainfall (mm)",
      na.value = NA,
      option = "C",
      limits = c(min_val, max_val)
    ) +
    theme_minimal() +
    labs(title = paste(month, "2021")) +
    theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
}

plots_full <- mapply(
  plot_month_full, months, raster_paths,
  MoreArgs = list(senegal = senegal, min_val = rainfall_min, max_val = rainfall_max),
  SIMPLIFY = FALSE
)

# Combine as 2x2
(plots_full[[1]] | plots_full[[2]]) /
(plots_full[[3]] | plots_full[[4]]) + plot_layout(guides = "collect") & theme(legend.position = 'bottom')

# Helper function for regions plot
plot_month_regions <- function(month, raster_path, senegal, roi_union, roi_sf, other_sf, min_val, max_val) {
  raster <- rast(raster_path)
  raster_senegal <- crop(raster, vect(senegal)) |> mask(vect(senegal))
  raster_roi <- mask(raster_senegal, vect(roi_union))
  ggplot() +
    geom_sf(data = other_sf, fill = "grey90", color = "grey", size = 0.2) +
    geom_spatraster(data = raster_roi, aes(fill = !!sym(paste0("senegal_precipitation_2021_", month, "_chirps"))), na.rm = TRUE) +
    geom_sf(data = roi_sf, fill = NA, color = "grey", size = 0.6) +
    # Optionally use geom_label_repel for better labels
    scale_fill_viridis_c(
      name = "Rainfall (mm)",
      na.value = NA,
      option = "C",
      limits = c(min_val, max_val)
    ) +
    theme_minimal() +
    labs(title = paste(month, "2021 (Regions)")) +
    theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
}

plots_regions <- mapply(
  plot_month_regions, months, raster_paths,
  MoreArgs = list(
    senegal = senegal,
    roi_union = roi_union,
    roi_sf = roi_sf,
    other_sf = other_sf,
    min_val = rainfall_min,
    max_val = rainfall_max
  ),
  SIMPLIFY = FALSE
)

# Combine as 2x2
(plots_regions[[1]] | plots_regions[[2]]) /
(plots_regions[[3]] | plots_regions[[4]]) + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
```