---
title: "Basic Statistics and Maps"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

## Summary

The goal of this page is to provide the map and a basic understanding of the data I have. This is to ensure visuals for publication have the same themeing but also to ensure that my data management aligns well with expectations and that there are no errors. If you see errors, please let me know!

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("basic_stats")

# Load gt and webshot2 for table export
library(gt)
library(webshot2)

# Helper function to save gt tables as PNG and CSV
save_gt_png <- function(gt_obj, filename, subfolder = "basic_stats") {
  # Create output directories
  png_dir <- here::here("outputs", "tables", subfolder, "png")
  csv_dir <- here::here("outputs", "tables", subfolder, "csv")
  dir.create(png_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Save PNG
  png_path <- file.path(png_dir, filename)
  gtsave(gt_obj, png_path, vwidth = 3000)
  
  # Save CSV - extract the underlying data from the gt object
  csv_filename <- sub("\\.png$", ".csv", filename)
  csv_path <- file.path(csv_dir, csv_filename)
  
  # Extract data from gt object
  gt_data <- gt_obj$`_data`
  write_csv(gt_data, csv_path)
  
  return(invisible(list(png = png_path, csv = csv_path)))
}
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false
#| results: hide

# Load and process data using the common function
data <- load_ose_data(cache_processed = TRUE)

# Save basic dataset summary
save_data_table(
  data |> 
    summarise(
      n_farmers = n_distinct(farmer),
      n_regions = n_distinct(region),
      n_missions = n_distinct(mission_number),
      total_observations = n(),
      mean_ose_count = mean(ose_count, na.rm = TRUE),
      mean_temperature = mean(temperature, na.rm = TRUE),
      mean_ground_cover = mean(percent_ground_cover, na.rm = TRUE),
      mean_yield = mean(rendement_en_kg_ha, na.rm = TRUE)
    ),
  "dataset_summary"
)
```

## Dataset Summary

```{r variable-summary}
#| echo: false
#| message: false
#| warning: false

# Basic statistics around dataset 
summary_table <- table_variable_summary(data)

# Display the summary table
summary_table
```

## Map of Senegal and Study Regions

```{r senegal-map}
#| echo: false
#| message: false
#| warning: false

senegal_map <- plot_senegal_map(
  regions_of_interest = STUDY_REGIONS,
  show_legend = FALSE,
  roi_color = "tomato",
  other_color = "grey80",
  label_size = 4
)

senegal_map
```

## Mean Annual Precipitation (MAP)

Mean annual precipitation (MAP) for Senegal was calculated using the CHIRPS dataset, which provides high-resolution (approximately 5 km) gridded precipitation estimates. We extracted CHIRPS daily precipitation data for Senegal for the most recent 10-year period (2015–2024). The data were spatially clipped to the national boundary, summed annually, and then averaged across years to produce a single raster layer representing mean annual precipitation.

```{r map-setup}
#| echo: false
#| message: false
#| warning: false

library(terra)
library(tidyterra)
library(sf)
library(rnaturalearth)
library(scales)

# Import the MAP TIFF
map_raster <- rast("data/processed/maps/senegal_map_2015_2024_chirps.tif")

# Get Senegal boundary and regions
senegal <- ne_states(country = "Senegal", returnclass = "sf")

# Define study regions
regions_of_interest <- c("Saint-Louis", "Thiès", "Fatick", "Kaffrine")
roi_sf <- senegal |> filter(name %in% regions_of_interest)
other_sf <- senegal |> filter(!name %in% regions_of_interest)

# Crop and mask raster to Senegal boundary
map_raster_senegal <- crop(map_raster, vect(senegal)) |>
  mask(vect(senegal))

# Create union of study regions for masking
roi_union <- st_union(roi_sf)
map_raster_roi <- mask(map_raster_senegal, vect(roi_union))
```

### Full Senegal Context

```{r}
library(terra)
library(sf)
library(smoothr)
library(ggplot2)

map_raster_senegal <- terra::disagg(map_raster_senegal, fact=4)

# 1. Smooth boundaries as usual
roi_sf_smooth     <- smoothr::smooth(roi_sf, method = "ksmooth", smoothness = 2)
senegal_sf_smooth <- smoothr::smooth(senegal, method = "ksmooth", smoothness = 2)

# 2. Masked raster (as before)
senegal_vect    <- vect(senegal)
raster_cropped  <- crop(map_raster_senegal, senegal_vect)
raster_masked   <- mask(raster_cropped, senegal_vect)
raster_masked[raster_masked == 0] <- NA
# 5. Plot (as polygons; use color=NA to avoid grid outlines; may try linetype=0)
ggplot() +
  geom_spatraster(data = raster_masked, 
                  aes(fill = senegal_map_2015_2024_chirps), 
                  na.rm = TRUE) +
  geom_sf(data = senegal_sf_smooth, fill = NA, color = "grey80", linewidth = 1.5) +
  geom_sf(data = roi_sf_smooth, fill = NA, color = 'gray85', linewidth = 2) +
  geom_sf(data = roi_sf_smooth, fill = NA, color = 'black',  linewidth = 1) +
  scale_fill_viridis_c(
      name      = "MAP (mm)", 
      na.value  = NA, 
      option    = "D", 
      direction = -1
  ) +
  theme_minimal() +
  labs(title = "Mean Annual Precipitation (2015–2024 averaged)") +
  theme(
    axis.text   = element_blank(),
    axis.ticks  = element_blank(),
    panel.grid  = element_blank()
  )
```

## 2021 Monthly Rainfall Patterns

Monthly rainfall patterns during the 2021 growing season (July-October) show the progression of precipitation across the study regions.

```{r monthly-rainfall-setup}
#| echo: false
#| message: false
#| warning: false

# Define months and raster file paths
months <- c("july", "august", "september", "october")
raster_paths <- paste0("data/processed/maps/senegal_precipitation_2021_", 
                       months, "_chirps.tif")

# Calculate global min/max across all months for consistent color scale
rainfall_values <- lapply(raster_paths, function(path) {
  r <- rast(path)
  values(r)[!is.na(values(r))]
})
all_values <- unlist(rainfall_values)
rainfall_min <- min(all_values)
rainfall_max <- max(all_values)
```

### Full Senegal Monthly Maps


```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10

library(terra)
library(sf)
library(smoothr)
library(ggplot2)
library(patchwork)
library(rlang)  # for !!sym() in aes

# Define months and raster file paths
months <- c("july", "august", "september", "october")
raster_paths <- paste0("data/processed/maps/senegal_precipitation_2021_", 
                       months, "_chirps.tif")

# Calculate global min/max across all months for consistent color scale
rainfall_values <- lapply(raster_paths, function(path) {
  r <- rast(path)
  values(r)[!is.na(values(r))]
})
all_values <- unlist(rainfall_values)
rainfall_min <- min(all_values)
rainfall_max <- max(all_values)

# --- Smooth the outlines ONCE outside loop, for speed ---
roi_sf_smooth      <- smoothr::smooth(roi_sf, method = "ksmooth", smoothness = 2)
senegal_sf_smooth  <- smoothr::smooth(senegal, method = "ksmooth", smoothness = 2)

# Helper function for full Senegal plot
plot_month_full <- function(month, raster_path, senegal, roi_sf,
                            senegal_sf_smooth, roi_sf_smooth, min_val, max_val) {
  raster <- rast(raster_path)
  # 1. Upsample (smooth raster-edged look)
  raster <- terra::disagg(raster, fact = 4)
  # 2. Mask to Senegal boundary
  senegal_vect <- vect(senegal)
  raster_cropped <- crop(raster, senegal_vect)
  raster_masked  <- mask(raster_cropped, senegal_vect)

  # -- Layered ggplot: smoothed outlines, no ugly squares, constant scale! --
  ggplot() +
    stars::geom_stars(data = stars::st_as_stars(raster_masked)) +
    geom_sf(data = senegal_sf_smooth, fill = NA, color = "grey80", linewidth = 1.5) +
    geom_sf(data = roi_sf_smooth, fill = NA, color = 'gray85', linewidth = 2) +
    geom_sf(data = roi_sf_smooth, fill = NA, color = 'black', linewidth = 1) +
    scale_fill_viridis_c(
      name = "Rainfall (mm)",
      na.value = NA,
      limits = c(min_val, max_val), 
      option = "D", 
      direction = -1
    ) +
    theme_minimal() +
    labs(title = paste(tools::toTitleCase(month), "2021")) +
    theme(axis.text = element_blank(), 
          axis.ticks = element_blank(), 
          panel.grid = element_blank()) +
    xlab('') +
    ylab('')
}

# Create plots
plots_full <- mapply(
  plot_month_full, months, raster_paths,
  MoreArgs = list(
    senegal = senegal, 
    roi_sf = roi_sf, 
    senegal_sf_smooth = senegal_sf_smooth, 
    roi_sf_smooth = roi_sf_smooth, 
    min_val = rainfall_min, 
    max_val = rainfall_max
  ),
  SIMPLIFY = FALSE
)

# Combine in 2x2 grid, grouped legend
(final_plot <- (plots_full[[1]] | plots_full[[2]]) /
  (plots_full[[3]] | plots_full[[4]]) + 
  patchwork::plot_layout(guides = "collect") & 
  theme(legend.position = "bottom"))
```

Below is the average monthly rainfall per region


```{r}
library(terra)
library(sf)
library(dplyr)

# Ensure roi_sf is in the same CRS as the rasters
roi_sf <- st_transform(roi_sf, crs(rast(raster_paths[1])))
# Function to extract mean and sd rainfall for each region in roi_sf for a given raster
extract_monthly_stats <- function(raster_path, polygons, month) {
  r <- rast(raster_path)
  # Extract mean
  mean_df <- terra::extract(r, vect(polygons), fun = mean, na.rm = TRUE)
  # Extract sd
  sd_df <- terra::extract(r, vect(polygons), fun = sd, na.rm = TRUE)
  # Combine and add region and month
  out <- mean_df |>
    rename(mean_rainfall = 2) |>
    left_join(sd_df |> rename(sd_rainfall = 2), by = "ID") |>
    mutate(region = polygons$name, month = month)
  out
}

# Loop over months and get stats
rainfall_stats <- lapply(seq_along(months), function(i) {
  extract_monthly_stats(raster_paths[i], roi_sf, months[i])
})

# Combine into one data frame
rainfall_stats_df <- bind_rows(rainfall_stats)

# Clean up columns if needed
rainfall_stats_df <- rainfall_stats_df |>
  select(region, month, mean_rainfall, sd_rainfall)


rainfall_stats_tbl <- rainfall_stats_df |>
  as_tibble() |>
  drop_na(mean_rainfall) |>
  select(region, month, mean_rainfall, sd_rainfall) |>
  mutate(
    mean_rainfall = round(mean_rainfall, 1),
    sd_rainfall = round(sd_rainfall, 1)
  ) |>
  gt(rowname_col = "region") |>
  tab_header(
    title = "Average Monthly Rainfall by Region (2021)"
  ) |>
  cols_label(
    region = "Region",
    month = "Month",
    mean_rainfall = "Mean Rainfall (mm)",
    sd_rainfall = "SD (mm)"
  ) |>
  fmt_number(
    columns = c(mean_rainfall, sd_rainfall),
    decimals = 1
  ) |>
  tab_options(
    table.font.size = "small"
  )

save_gt_png(rainfall_stats_tbl, "average_monthly_rainfall_by_region.png")
rainfall_stats_tbl

```