---
title: "Basic Statistics and Maps"
format: html
toc: true
---

## Summary

The goal of this page is to provide the map and a basic understanding of the data I have. This is to ensure visuals for publication have the same themeing but also to ensure that my data management aligns well with expectations and that there are no errors. If you see errors, please let me know!

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("basic_stats")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false
#| results: hide

# Load and process data using the common function
data <- load_ose_data(cache_processed = TRUE)

# Save basic dataset summary
save_data_table(
  data |> 
    summarise(
      n_farmers = n_distinct(farmer),
      n_regions = n_distinct(region),
      n_missions = n_distinct(mission_number),
      total_observations = n(),
      mean_ose_count = mean(ose_count, na.rm = TRUE),
      mean_temperature = mean(temperature, na.rm = TRUE),
      mean_ground_cover = mean(percent_ground_cover, na.rm = TRUE),
      mean_yield = mean(rendement_en_kg_ha, na.rm = TRUE)
    ),
  "dataset_summary"
)
```

## Dataset Summary

```{r variable-summary}
#| echo: false
#| message: false
#| warning: false

# Basic statistics around dataset
summary_table <- table_variable_summary(data)

# Also save the detailed variable summary
if (exists("summary_table")) {
  # Extract the underlying data if it's a gt table
  tryCatch({
    if ("gt_tbl" %in% class(summary_table)) {
      # For gt tables, we need to recreate the summary data
      summary_data <- data |>
        select(where(is.numeric)) |>
        summarise(across(everything(), list(
          min = ~min(.x, na.rm = TRUE),
          max = ~max(.x, na.rm = TRUE), 
          mean = ~mean(.x, na.rm = TRUE),
          median = ~median(.x, na.rm = TRUE),
          sd = ~sd(.x, na.rm = TRUE)
        ))) |>
        pivot_longer(everything(), names_to = "variable_stat", values_to = "value") |>
        separate(variable_stat, into = c("variable", "statistic"), sep = "_(?=[^_]*$)")
      
      save_data_table(summary_data, "variable_summary_statistics")
    }
  }, error = function(e) {
    message("Could not save summary table data: ", e$message)
  })
}

# Display the summary table
summary_table
```

## Map of Senegal and Study Regions

```{r senegal-map}
#| echo: false
#| message: false
#| warning: false

senegal_map <- plot_senegal_map(
  regions_of_interest = STUDY_REGIONS,
  show_legend = FALSE,
  roi_color = "tomato",
  other_color = "grey80",
  label_size = 4
)

senegal_map
```

## Mean Annual Precipitation (MAP) 

Mean annual precipitation (MAP) for Senegal was calculated using the CHIRPS dataset, which provides high-resolution (approximately 5 km) gridded precipitation estimates. We extracted CHIRPS daily precipitation data for Senegal for the most recent 10-year period (2015–2024). The data were spatially clipped to the national boundary, summed annually, and then averaged across years to produce a single raster layer representing mean annual precipitation.

```{r map-setup}
#| echo: false
#| message: false
#| warning: false

library(terra)
library(tidyterra)
library(sf)
library(rnaturalearth)
library(scales)

# Import the MAP TIFF
map_raster <- rast("data/processed/maps/senegal_map_2015_2024_chirps.tif")

# Get Senegal boundary and regions
senegal <- ne_states(country = "Senegal", returnclass = "sf")

# Define study regions
regions_of_interest <- c("Saint-Louis", "Thiès", "Fatick", "Kaffrine")
roi_sf <- senegal |> filter(name %in% regions_of_interest)
other_sf <- senegal |> filter(!name %in% regions_of_interest)

# Crop and mask raster to Senegal boundary
map_raster_senegal <- crop(map_raster, vect(senegal)) |>
  mask(vect(senegal))

# Create union of study regions for masking
roi_union <- st_union(roi_sf)
map_raster_roi <- mask(map_raster_senegal, vect(roi_union))
```

### Option 1: Full Senegal Context

```{r map-full-senegal}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 6
ggplot() +
  geom_sf(data = senegal) +
  geom_spatraster(data = map_raster_senegal, 
                  aes(fill = senegal_map_2015_2024_chirps), 
                  na.rm = TRUE) +
  geom_sf(data = senegal, fill = NA, color = "grey", linewidth = 0.6) +
  scale_fill_viridis_c(name = "MAP (mm)", na.value = NA, option = "C") +
  theme_minimal() +
  labs(title = "Mean Annual Precipitation (2015–2024)",
       subtitle = "Full Senegal with study regions visible") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

### Option 2: Study Regions Only

```{r map-regions-only}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 6

ggplot() +
  geom_sf(data = other_sf, fill = "grey90", color = "grey", linewidth = 0.2) +
  geom_spatraster(data = map_raster_roi, 
                  aes(fill = senegal_map_2015_2024_chirps), 
                  na.rm = TRUE) +
  geom_sf(data = roi_sf, fill = NA, color = "grey", linewidth = 0.6) +
  geom_sf_text(data = roi_sf, aes(label = name), 
               size = 4, fontface = "bold", color = "black") +
  scale_fill_viridis_c(name = "MAP (mm)", na.value = NA, option = "C") +
  theme_minimal() +
  labs(title = "Mean Annual Precipitation (2015–2024)",
       subtitle = "Study regions with labels") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

### Option 3: Study Regions with Mean Values

```{r map-regions-with-values}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 6

# Extract mean MAP values for each region
roi_sf$mean_map <- terra::extract(map_raster, vect(roi_sf), 
                                   fun = mean, na.rm = TRUE)[, 2]

# Create label with region name and mean MAP value
roi_sf$label <- sprintf(
  "%s\n(%s mm)",
  roi_sf$name,
  comma(round(roi_sf$mean_map, 0))
)

# Define color palette for regions
region_palette <- c(
  "Saint-Louis" = "#1b9e77",
  "Thiès"       = "#d95f02",
  "Fatick"      = "#7570b3",
  "Kaffrine"    = "#e7298a"
)

# Create map with colored regions and mean precipitation values
ggplot() +
  geom_sf(data = other_sf, fill = "grey90", color = "grey", linewidth = 0.2) +
  geom_sf(data = roi_sf, aes(fill = name), color = "black", linewidth = 0.6) +
  geom_sf_text(
    data = roi_sf,
    aes(label = label),
    size = 4.5, 
    fontface = "bold",
    color = "white",
    bg.color = "black",
    bg.r = 0.15
  ) +
  scale_fill_manual(
    name = NULL,
    values = region_palette,
    guide = "none"
  ) +
  theme_minimal() +
  labs(
    title = "Mean Annual Precipitation by Region (2015–2024)",
    subtitle = "Study regions with mean precipitation values"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

## 2021 Monthly Rainfall Patterns

Monthly rainfall patterns during the 2021 growing season (July-October) show the progression of precipitation across the study regions.

```{r monthly-rainfall-setup}
#| echo: false
#| message: false
#| warning: false

# Define months and raster file paths
months <- c("july", "august", "september", "october")
raster_paths <- paste0("data/processed/maps/senegal_precipitation_2021_", 
                       months, "_chirps.tif")

# Calculate global min/max across all months for consistent color scale
rainfall_values <- lapply(raster_paths, function(path) {
  r <- rast(path)
  values(r)[!is.na(values(r))]
})
all_values <- unlist(rainfall_values)
rainfall_min <- min(all_values)
rainfall_max <- max(all_values)
```

### Full Senegal Monthly Maps

```{r monthly-rainfall-full}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10

# Helper function for full Senegal plot
plot_month_full <- function(month, raster_path, senegal, min_val, max_val) {
  raster <- rast(raster_path)
  raster_senegal <- crop(raster, vect(senegal)) |> 
    mask(vect(senegal))
  
  ggplot() +
    geom_sf(data = senegal) +
    geom_spatraster(data = raster_senegal, 
                    aes(fill = !!sym(paste0("senegal_precipitation_2021_", month, "_chirps"))), 
                    na.rm = TRUE) +
    geom_sf(data = senegal, fill = NA, color = "grey", linewidth = 0.6) +
    scale_fill_viridis_c(
      name = "Rainfall (mm)",
      na.value = NA,
      option = "C",
      limits = c(min_val, max_val)
    ) +
    theme_minimal() +
    labs(title = paste(tools::toTitleCase(month), "2021")) +
    theme(axis.text = element_blank(), 
          axis.ticks = element_blank(), 
          panel.grid = element_blank())
}

plots_full <- mapply(
  plot_month_full, months, raster_paths,
  MoreArgs = list(senegal = senegal, min_val = rainfall_min, max_val = rainfall_max),
  SIMPLIFY = FALSE
)

# Combine plots in 2x2 grid
(plots_full[[1]] | plots_full[[2]]) /
(plots_full[[3]] | plots_full[[4]]) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```

### Study Regions Monthly Maps

```{r monthly-rainfall-regions}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10

# Helper function for regions plot
plot_month_regions <- function(month, raster_path, senegal, roi_union, roi_sf, other_sf, min_val, max_val) {
  raster <- rast(raster_path)
  raster_senegal <- crop(raster, vect(senegal)) |> 
    mask(vect(senegal))
  raster_roi <- mask(raster_senegal, vect(roi_union))
  
  ggplot() +
    geom_sf(data = other_sf, fill = "grey90", color = "grey", linewidth = 0.2) +
    geom_spatraster(data = raster_roi, 
                    aes(fill = !!sym(paste0("senegal_precipitation_2021_", month, "_chirps"))), 
                    na.rm = TRUE) +
    geom_sf(data = roi_sf, fill = NA, color = "grey", linewidth = 0.6) +
    scale_fill_viridis_c(
      name = "Rainfall (mm)",
      na.value = NA,
      option = "C",
      limits = c(min_val, max_val)
    ) +
    theme_minimal() +
    labs(title = paste(tools::toTitleCase(month), "2021")) +
    theme(axis.text = element_blank(), 
          axis.ticks = element_blank(), 
          panel.grid = element_blank())
}

plots_regions <- mapply(
  plot_month_regions, months, raster_paths,
  MoreArgs = list(
    senegal = senegal,
    roi_union = roi_union,
    roi_sf = roi_sf,
    other_sf = other_sf,
    min_val = rainfall_min,
    max_val = rainfall_max
  ),
  SIMPLIFY = FALSE
)

# Combine plots in 2x2 grid
(plots_regions[[1]] | plots_regions[[2]]) /
(plots_regions[[3]] | plots_regions[[4]]) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```