---
title: "2021 Farmer gender analysis"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

## Summary

The goal of this page is to analyze the difference between OSE count, damage and millet between farmer gender.

TLDR:

-   no significant impact of gender on OSE count

-   Male farmers had lower locust damage than female farmers

-   Female farmers had lower yield than male farmers

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Load processed data
data <- load_ose_data(cache_processed = TRUE)

data <- data |>
    mutate(ose_damage_percent = ose_damage_percent/100,mission_number=factor(mission_number)) |>
    select(year,region,mission_number,farmer_gender,ose_count,ose_damage_percent,rendement_en_kg_ha)
```

# OSE count analysis

There doesnt seem to be a big difference between farmer gender and density. A poisson model was insigificant.

```{r}

data |>
    ggplot(aes(x=farmer_gender,y=ose_count)) +
        geom_boxplot(outlier.size=-1) +
        geom_jitter(width=0.2,height=0) +
        facet_wrap(~mission_number,ncol=1)

```

I will create a quick poisson glmm with mission_number as a random effect and farmer gender as a fixed effect to see if there is a difference between male and female farmer.

```{r}

density_mod <- glmmTMB(ose_count ~ farmer_gender + (1|region) + (1|mission_number),data=data,
    family=tweedie())

summary(density_mod)

resids <- simulateResiduals(density_mod)
plot(resids)

```

```{r}

emm <- emmeans(density_mod,~farmer_gender,type='response') |> as_tibble()


emm |> 
    ggplot(aes(x=farmer_gender,y=response)) +
        geom_point(size=5) +
        geom_errorbar(aes(ymin=asymp.LCL,ymax=asymp.UCL),width=0) +
        ylab('OSE count (marginal means)') +
        xlab("")

```

# OSE damage

```{r}

data |>
    ggplot(aes(x=farmer_gender,y=ose_damage_percent)) +
        geom_boxplot(outlier.size=-1) +
        geom_jitter(width=0.2,height=0) +
        facet_wrap(~mission_number,ncol=1)

```

I will create a quick poisson glmm with mission_number as a random effect and farmer gender as a fixed effect to see if there is a statistical significance.

```{r}

damage_mod <- glmmTMB(ose_damage_percent ~ farmer_gender + (1|region) + (1|mission_number),data=data,
    family=tweedie())

summary(damage_mod)

resids <- simulateResiduals(damage_mod)
plot(resids)

```

```{r}

emm <- emmeans(damage_mod,~farmer_gender,type='response') |> as_tibble()


emm |> 
    ggplot(aes(x=farmer_gender,y=response)) +
        geom_point(size=5) +
        geom_errorbar(aes(ymin=asymp.LCL,ymax=asymp.UCL),width=0) +
        ylab('OSE count (marginal means)') +
        xlab("")

```

# Millet Yield

```{r}

yield_dat <- data |>
    drop_na(rendement_en_kg_ha)

yield_dat |>
    ggplot(aes(x=farmer_gender,y=rendement_en_kg_ha)) +
        geom_boxplot(outlier.size=-1) +
        geom_jitter(width=0.2,height=0) 

```

```{r}

yield_mod <- glmmTMB(rendement_en_kg_ha ~ farmer_gender + (1|region) + (1|mission_number),data=yield_dat,family=tweedie())

summary(yield_mod)

resids <- simulateResiduals(yield_mod)
plot(resids)

```

```{r}

emm <- emmeans(yield_mod,~farmer_gender,type='response') |> as_tibble()


emm |> 
    ggplot(aes(x=farmer_gender,y=response)) +
        geom_point(size=5) +
        geom_errorbar(aes(ymin=asymp.LCL,ymax=asymp.UCL),width=0) +
        ylab('OSE count (marginal means)') +
        xlab("")

```