---
title: "2021 Locust Density Analysis"
subtitle: "figures and tables"
format: html
toc: true
---

## Summary

This page presents the visualizations and statistical analyses conducted on the 2021 field survey dataset, supporting ongoing manuscript development for publication.

The analyses focus on Senegalese locust (OSE) density and ground cover data collected across multiple regions and fertilizer treatments in Senegalese agricultural fields. The goal of this page is to replicate the current figures in the manuscript. This is to ensure visuals are made in the same way but also to ensure that my data management aligns well with expectations and that there are no errors. If you see errors, please let me know!

## Major questions:

1.  OSE densiy plots: do you prefer option 1 or 2 (assuming that we could workshop either one to be better)?
2.  OSE densiy x ground cover: I am unable to replicate the curves seen in the mansucript (figure 5), can you confirm how the data was handled in this analysis?
3.  annual OSE density: my averages do not add up to the numbers reported in table 1 in the manuscript, can you confirm how the data was handled for this table?
4.  For model summaries (e.g. p values) how would we like to handle them? Would you prefer in-line references or supplementary tables?


```{r package loading}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false


library(tidyverse)
library(rnaturalearth)
library(janitor)
library(here)
library(patchwork)
library(ggpubr)
library(gt)
library(glmmTMB)
library(mgcv)
library(gratia)
library(emmeans)

here::i_am('README.md')

# setting project root
# sourcing functions
source(('R/functions/data_preprocessing.R'), local = knitr::knit_global())
source(('R/functions/table_variable_summary.R'), local = knitr::knit_global())
source(('R/functions/senegal_map.R'), local = knitr::knit_global())
source(('R/functions/locust_density_modeling.R'), local = knitr::knit_global())

```

```{r data loading and processing}
#| echo: false
#| message: false
#| warning: false

data = process_senegal_data('data/raw/Toure_OSE2021data_v03.xlsx')

```

# Basic statistics around dataset

```{r basic_statistics}
#| echo: false
#| message: false
#| warning: false


table_variable_summary(data)

```

# Map of senegal and regions

```{r senegal_map}
#| echo: false
#| message: false
#| warning: false


plot_senegal_map(
  regions_of_interest = c('Saint-Louis', 'Thiès', 'Fatick', 'Kaffrine'),
  show_legend = FALSE,
  roi_color = "tomato",
  other_color = "grey80",
  label_size = 4)

```

# Locust density and region by treatment

To assess how locust count changes between treatment (fertilized and control) and region (five regions), I built a generalized linear mixed model (family: Poisson) per mission as follows:

\$\$

$$ \begin{align*}
\text{ose count}_{ij} =\ 
  &\beta_0 \\
  &+ \beta_1 \cdot \text{fertilizer treatment}_i \\
  &+ \beta_2 \cdot \text{region}_i \\
  &+ \beta_3 \cdot (\text{fertilizer treatment}_i \times \text{region}_i) \\
  &+ u_j \\
  &+ \varepsilon_{ij}
\end{align*}ilon\_{ij} \\end{align\*} $$

\$\$

Where:\
- $\text{ose count}_{ij}$: observed ose count for observation $i$ from farmer $j$\
- $\text{fertilizer treatment}_i$: fertilizer treatment group of observation $i$\
- $\text{region}_i$: region of observation $i$\
- $(\text{fertilizer treatment}_i \times \text{region}_i)$: interaction between fertilizer treatment and region\
- $u_j$: random effect for farmer $j$, $u_j \sim N(0, \sigma^2_{farmer})$\
- $\varepsilon_{ij}$: residual error

Below are the diagnostic plots for these models:

```{r computing ose density models}
#| echo: false
#| message: false
#| warning: false

# Split data into a list of data based on mission date
model_data <- data |>
    group_split(mission_number)

#create lists to store objects
mission_mods <- list()
mission_diagnostic_plots <- list()
mission_emmeans <- list()

#seek through lists. create model, diagnostic plot, and emmeans
for (i in seq_along(model_data)){
    diagnostic_plot_title = paste0('mission',i)
    mission_mods[[i]] <- create_glmm_model(model_data[[i]])
    mission_diagnostic_plots[[i]] <- plot_model_diagnostics(mission_mods[[i]],plot_title = diagnostic_plot_title)
    mission_emmeans[[i]] <- get_emmeans_and_pairs(mission_mods[[i]], mission_num = i)
}

names(mission_mods) <- c('m1','m2','m3')
names(mission_diagnostic_plots) <- c('m1','m2','m3')
names(mission_emmeans) <- c('m1','m2','m3')
```

```{r plot diagnostic plots}

diagnostic_plots <- (mission_diagnostic_plots$m1 +
    mission_diagnostic_plots$m2) /
    (mission_diagnostic_plots$m3 +
    plot_spacer())

diagnostic_plots
```

These models are not perfect, but I think they are fine to use for this analysis. Happy to invetigate further if you would like me too.

## OSE density by treatment and region (option 1)

```{r densiy option 1 plot}
#| out-width: 100%
#| out-height: 100%
#| 
# A list of model results with emmeans
mission_emms <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    # Add the list element name as a column:
    df <- mission_emmeans[[nm]]$emmeans
    df$source <- nm
    df
  }),
  .id = 'mission_numer' 
)

count_plot_option_1 <- plot_mission_density(
    data |> drop_na(ose_count,fertilizer_treatment),
    mission_emms,
    regions = c('Saint Louis','Thies','Fatick','Kaffrine'),
    legend_position = 'bottom'
) 

count_plot_option_1

```

## OSE density by treatment and region (option 2)

Instead of a plot per mission, lets plot the missions together (x-axis) and see how region counts shifted through time

```{r count plot option 2}

count_plot_option_2 <- plot_emmeans(mission_emms,
    color_palette = "Degas",
    ylab = "OSE count",
    xlab = "Mission")

count_plot_option_2

```

## OSE density by teatment and region model summaries (tabulated)

Here are the model summaries and pairwise comparisons. This will be *across* missions, but please keep in mind each mission had it's own model. So we cant make across-mission comparions just yet.

```{r}
library(broom.mixed)
mission_mod_summary <- bind_rows(
  lapply(names(mission_mods), function(nm) {
    # Add the list element name as a column:
    df <- mission_mods[[nm]] |> tidy()
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

mission_emms <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    # Add the list element name as a column:
    df <- mission_emmeans[[nm]]$emmeans
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

mission_pairs <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    # Add the list element name as a column:
    df <- mission_emmeans[[nm]]$pairwise
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

library(gt)
library(dplyr)


mission_mod_summary <- mission_mod_summary |>
  arrange(mission_number) |>
  mutate(effect = ifelse(effect == 'ran_pars', 'random', 'fixed'),
    term = ifelse(term == 'sd__(Intercept)','farmer',term),
    term = gsub('fertilizer_treatment|region','',term)) |>
  select(!c(component,group))



mission_mod_summary |>
  select(!source) |>
  gt(groupname_col = "mission_number") |>
  fmt_number(
    columns = c(estimate, std.error, statistic, p.value),
    decimals = 3
  ) |>
  cols_label(
    mission_number = "Mission",
    effect = "Effect",
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. Error",
    statistic = "Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "OSE density by treatment and region by mission"
  ) |>
  as_raw_html()

```

Here are the pairwise comparisons between region and treatment by mission number. Fair warning, its a large table since there are a lot of regions.

```{r}

mission_pairs |>
  arrange(contrast) |>
  select(!source) |>
  gt(groupname_col = "mission_number") |>
  fmt_number(
    columns = c(ratio, SE, z.ratio, p.value),
    decimals = 3
  ) |>
  tab_header(
    title = "OSE density model pairwise comparisons by mission"
  ) |>
  as_raw_html()

```

# OSE density x plant cover

We would like to see how density is influenced by plant cover. To do this I am building a generalized additive model (family: tweedie). Following this form:

The generalized additive model (GAM) fitted is:

\$\$

$$ \begin{align*}
\text{ose\_count}_i = \\
  &\alpha + \\ 
  &+ f(\text{percent\_ground\_cover}_i,\, \text{region\_treat}_i) +  \\
  &+ b_{\text{farmer}_i} + \\
  &+ \varepsilon_i
\end{align*} $$

\$\$

where: - $\alpha$ is the intercept, - $f(\text{percent\_ground\_cover},\, \text{region\_treat})$ is a smooth function (tensor product, using factor-smooth interaction) over percent ground cover and region treatment, - $b_{\text{farmer}_i}$ is a random effect for each farmer, - $\varepsilon_i$ is the residual error.

```{r constructing gam model}
#| echo: false
#| message: false
#| warning: false


plant_cover_dat <- data |>
    dplyr::select(region,farmer,fertilizer_treatment,mission_number,ose_count,percent_ground_cover) |>
    mutate(
        region_treat = factor(paste0(region,'_',fertilizer_treatment))
    ) |>
    drop_na(ose_count,fertilizer_treatment)

plant_cover_mod <- build_gam(plant_cover_dat,
    formula = ose_count ~ te(percent_ground_cover, region_treat, bs='fs', k=10) + s(farmer, bs='re'),
    nthreads = 9,
    family = mgcv::tw(),
    k = 10) 

```

The model diagnostics look pretty good! I tried to build a poisson model, but the data was overdispersed. Instead I went with a tweedie model which does a pretty good job at handling this type of data:

```{r gam model diagnostics}

appraise(plant_cover_mod)

```

However, I am not able to reproduce the smooths seen in the manuscript sent to me. This is likely due to data differences as I am not sure what was included in the initial modeling approach. **Roughly** we can see similiar trends in Kaffrine, Fatick, and Saint Louis. But Thies definitely has a different trend. Additionally, there seems to be different trends between control and fertilized treatments as comapred to the current plots in the manuscript. Not sure what to think here.

```{r plotting gam model results}
#| echo: false
#| message: false
#| warning: false

gam_plot <- plot_gam_smooths_gratia(plant_cover_mod)

gam_plot
```

## GAM model summary

```{r}

broom.mixed::tidy(plant_cover_mod) |>
  gt() |>
  fmt_number(
    columns = c(edf, ref.df, statistic, p.value),
    decimals = 2
  ) |>
  cols_label(
    term = "Term",
    edf = "Estimated DF",
    ref.df = "Reference DF",
    statistic = "F Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "GAM model Summary: OSE Density x Plant Cover"
  ) |>
  tab_options(
    table.font.size = "small",
    row.striping.background_color = "#F6F6EF"
  ) |>
  as_raw_html()


```

# Table: annual OSE density by region and field type

The data discrepancies seen the the OSE count x ground cover are really shown in this table as well. We can see that the trends are the same (e.g. fertilized plots have less than control), but the actual averages are totally different.

```{r}

ose_dat <- data |>
  select(region, fertilizer_treatment, ose_count) |>
  mutate(region = factor(region, levels = c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')),
         fertilizer_treatment = factor(fertilizer_treatment, levels = c('fertilized', 'control'))) |>
  drop_na(ose_count, fertilizer_treatment) |>
  group_by(region, fertilizer_treatment) |>
  summarize(ose_mean = mean(ose_count, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = region, values_from = ose_mean)

# Create the gt table
ose_dat |>
  gt(rowname_col = "fertilizer_treatment") |>
  tab_header(
    title = "Mean OSE Count by Region and Fertilizer Treatment"
  ) |>
  fmt_number(
    columns = everything(),
    decimals = 2
  ) |>
  cols_label(
    fertilizer_treatment = "Treatment",
    `Saint Louis` = "Saint Louis",
    Thies = "Thies",
    Fatick = "Fatick",
    Kaffrine = "Kaffrine"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )


```