---
title: "2021 Locust × Ground Cover"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("locust_density_ground_cover")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data
data <- load_ose_data(cache_processed = TRUE)

```

## Overview

This analysis explores how locust (OSE) density is influenced by ground cover across different regions and fertilizer treatments. We use generalized additive models (GAMs) with a Tweedie family distribution to handle the overdispersed count data.

------------------------------------------------------------------------

# Analysis 1: Region-Treatment Interaction Model (Initial Approach)

This initial model uses a tensor product smooth to examine how the relationship between ground cover and OSE count varies across region-treatment combinations.

## Model Specification

$$
\begin{align*}
\text{ose\_count}_i &= \alpha + f(\text{percent\_ground\_cover}_i, \text{region\_treat}_i) + b_{\text{farmer}_i} + \varepsilon_i
\end{align*}
$$

where:

-   $\alpha$ is the intercept,
-   $f(\text{percent\_ground\_cover},\, \text{region\_treat})$ is a smooth function (tensor product, using factor-smooth interaction) over percent ground cover and region treatment,
-   $b_{\text{farmer}_i}$ is a random effect for each farmer,
-   $\varepsilon_i$ is the residual error.

## Model Fitting

```{r region-treatment-model}
#| echo: false
#| message: false
#| warning: false

# Prepare data for region-treatment interaction model
ground_cover_dat <- data |>
    select(region, farmer, fertilizer_treatment, mission_number, 
           ose_count, percent_ground_cover) |>
    mutate(
        region_treat = factor(paste0(region, '_', fertilizer_treatment)),
        farmer = factor(farmer)
    ) |>
    drop_na(ose_count, fertilizer_treatment)

# Use model caching for the GAM (can take several minutes to fit)
ground_cover_mod <- cache_model(
    model_name = "locust_ground_cover_gam",
    data = ground_cover_dat,
    fit_function = function(data, formula, nthreads, family) {
        build_gam(data, 
                 formula = formula, 
                 nthreads = nthreads, 
                 family = family)
    },
    # Simpler model for CI environments (faster fitting)
    fit_function_simple = function(data, formula, nthreads, family) {
        # Use simpler GAM with reduced complexity for CI
        build_gam(data,
                 formula = ose_count ~ te(percent_ground_cover, region_treat, bs = 'fs', k = 5) + s(farmer, bs = 're') +
                 mission_number,
                 nthreads = 1,
                 family = family)
    },
    formula = ose_count ~ te(percent_ground_cover, region_treat, bs = 'fs', k = 10) + s(farmer, bs = 're'),
    nthreads = 4,  # Reduced from 9 for better compatibility
    family = mgcv::tw()
)
```

The model diagnostics look pretty good! I tried to build a poisson model, but the data was overdispersed. Instead I went with a tweedie model which does a pretty good job at handling this type of data:

```{r gam model diagnostics}
#| echo: false
#| message: false
#| warning: false


appraise(ground_cover_mod)
```

## Model Results

The smoothed relationships show regional variation in how ground cover affects OSE density. Similar trends are observed in Kaffrine, Fatick, and Saint-Louis, while Thiès shows a distinct pattern. The control and fertilized treatments also exhibit different relationships.

**Note:** These results may differ from earlier manuscript versions due to data processing differences.

```{r region-treatment-results}
#| echo: false
#| message: false
#| warning: false

gam_plot <- plot_gam_smooths_gratia(ground_cover_mod)

gam_plot
```

## Model Summary Statistics

```{r region-treatment-summary}
#| echo: false
#| message: false
#| warning: false

broom.mixed::tidy(ground_cover_mod) |>
  gt() |>
  fmt_number(
    columns = c(edf, ref.df, statistic, p.value),
    decimals = 2
  ) |>
  cols_label(
    term = "Term",
    edf = "Estimated DF",
    ref.df = "Reference DF",
    statistic = "F Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "GAM Model Summary: OSE Density × Ground Cover (Region-Treatment Model)"
  ) |>
  tab_options(
    table.font.size = "small",
    row.striping.background_color = "#F6F6EF"
  ) |>
  as_raw_html()


```

------------------------------------------------------------------------

# Analysis 2: Additive Random Effects Model (Current Analysis)

*Updated: December 26, 2025*

This refined approach uses a simpler model structure with additive random effects rather than complex interactions. This model includes:

-   A smooth term for percent ground cover
-   Random effects for region, fertilizer treatment, mission number, and farmer

Key findings:

-   The relationship between ground cover and OSE count is non-linear, peaking around 50% ground cover
-   OSE counts are lower at both low and high ground cover
-   The model explains approximately 52% of deviance in the data

## Model Specification and Fitting

```{r additive-model-fit}
#| echo: false
#| message: false
#| warning: false


# Prepare data with all factors
data_filtered <- data |>
    select(year, region, farmer, farmer_gender, fertilizer_treatment, 
           mission_number, ose_count, percent_ground_cover) |>
    mutate(
      fertilizer_treatment = factor(fertilizer_treatment),
      region = factor(region),
      farmer = factor(farmer),
      mission_number = factor(mission_number)
    )

# Fit additive model with random effects
mod <- bam(
  ose_count ~ 
    s(percent_ground_cover) + 
    s(region, bs = "re") +
    s(fertilizer_treatment, bs = "re") +
    s(mission_number, bs = "re") +
    s(farmer, bs = "re"),
  data = data_filtered,
  family = tw()
)

# Model diagnostics
gratia::appraise(mod)

```

## Ground Cover Effect

The smooth effect of ground cover on OSE count shows a clear non-linear relationship:

```{r additive-model-ground-cover-effect}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Extract smooth estimates for ground cover
est <- smooth_estimates(mod, select = "percent_ground_cover", partial_match = TRUE)

# Adjust estimates to response scale
est$adj_est <- est$.estimate + coef(mod)[1]

# Plot the relationship
ggplot(est, aes(x = percent_ground_cover, y = adj_est)) +
  geom_ribbon(aes(ymin = adj_est - .se, ymax = adj_est + .se), 
              alpha = 0.2, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    x = "Ground Cover (%)",
    y = "Estimated OSE Count",
    title = "Effect of Ground Cover on OSE Density"
  ) +
  theme_minimal()
```

## Ground Cover Patterns

### Ground Cover by Mission

Ground cover increased throughout the growing season across all regions, with minimal regional differences.

```{r ground-cover-mission}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

data_filtered |>
  ggplot(aes(x = mission_number, y = percent_ground_cover)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.4) +
  facet_wrap(~region) +
  labs(
    title = "Ground Cover Progression Through Growing Season",
    x = "Mission Number",
    y = "Ground Cover (%)"
  ) +
  theme_minimal()
```

### Ground Cover by Fertilizer Treatment

Ground cover did not differ significantly between fertilizer treatments, but did vary between missions. No significant interaction was detected.

```{r ground-cover-fertilizer-treatment}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

data_filtered |>
  drop_na(percent_ground_cover, fertilizer_treatment) |>
  ggplot(aes(x = mission_number, y = percent_ground_cover, 
             fill = fertilizer_treatment)) +
  geom_boxplot(
    position = position_dodge(width = 0.8),
    outlier.shape = NA
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.4, jitter.height = 0),
    alpha = 0.5,
    size = 0.8
  ) +
  facet_wrap(~region) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Ground Cover by Fertilizer Treatment and Mission",
    x = "Mission Number",
    y = "Ground Cover (%)",
    fill = "Treatment"
  ) +
  theme_minimal()

# Statistical test for fertilizer × mission interaction
ground_cover_test <- glmmTMB(
  percent_ground_cover ~ fertilizer_treatment * mission_number,
  data = data_filtered
)
summary(ground_cover_test)
```

------------------------------------------------------------------------

# Analysis 3: Fertilizer-Specific Ground Cover Effects

*Model Comparison Analysis*

To determine whether the effect of ground cover on OSE count varies by fertilizer treatment, we compare multiple model structures using AIC.

## Model Comparison

```{r fertilizer-interaction-models}
#| echo: false
#| message: false
#| warning: false

# Null model (baseline)
null_mod <- bam(
  ose_count ~ 
    1 +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Model with fertilizer-specific smooth for ground cover
factor_smooth_mod <- bam(
  ose_count ~ 
    te(percent_ground_cover, by = fertilizer_treatment, k = 30) +
    fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Linear interaction model
linear_mod <- bam(
  ose_count ~ 
    percent_ground_cover * fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Single smooth without fertilizer interaction
without_fertilizer_mod <- bam(
  ose_count ~ 
    te(percent_ground_cover, k = 30) +
    fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Compare models using AIC


aic_tbl <- AIC(
  null_mod,
  factor_smooth_mod,
  without_fertilizer_mod,
  linear_mod
) |>
  as_tibble(rownames = "model") |>
  mutate(
    model_name = case_when(
      model == "null_mod" ~ "Null Model",
      model == "factor_smooth_mod" ~ "Smooth by Fertilizer Treatment",
      model == "without_fertilizer_mod" ~ "No Fertilizer Treatment Smooth",
      model == "linear_mod" ~ "Linear Interaction",
      TRUE ~ model
    )
  ) |>
  mutate(delta_AIC = AIC - min(AIC)) |>
  arrange(delta_AIC) |>
  select(model_name, AIC, delta_AIC)

aic_tbl
```

## Model Selection Results

The model with fertilizer-specific smooths ("Smooth by Fertilizer Treatment") was strongly supported by AIC (Δ AIC = 0). This indicates that fertilizer treatment modifies the functional relationship between ground cover and OSE count.

## Fertilizer-Specific Effects

```{r fertilizer-specific-smooths}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Extract smooth estimates for each fertilizer treatment
ests <- gratia::smooth_estimates(factor_smooth_mod, 
                                 select = 'percent', 
                                 partial = TRUE)

# Adjust to response scale
ests$adj_est <- ests$.estimate + coef(factor_smooth_mod)[1]

# Plot treatment-specific relationships
ests |>
  ggplot(aes(x = percent_ground_cover)) +
  geom_ribbon(
    aes(ymin = adj_est - .se, ymax = adj_est + .se, 
        fill = fertilizer_treatment),
    alpha = 0.2
  ) +
  geom_line(
    aes(y = adj_est, color = fertilizer_treatment),
    linewidth = 1.2
  ) +
  labs(
    x = "Ground Cover (%)",
    y = "Estimated OSE Count",
    title = "Ground Cover Effects by Fertilizer Treatment",
    color = "Treatment",
    fill = "Treatment"
  ) +
  scale_color_met_d(name = "Demuth") +
  scale_fill_met_d(name = "Demuth") +
  theme_minimal()


```

## Interpretation

### Key Findings

1.  **Model Selection**: The fertilizer-specific smooth model was strongly preferred (Δ AIC = 0), indicating that fertilizer treatment fundamentally alters the ground cover–OSE density relationship.

2.  **Peak Densities**:

    -   **Fertilized plots**: OSE counts peak at approximately 40% ground cover
    -   **Control plots**: OSE counts peak at approximately 60–70% ground cover

3.  **Mechanism**: Fertilizer treatment likely modifies OSE responses through:

    -   Changes in crop vigor and nutritional quality
    -   Altered plant species composition
    -   Modified microclimate conditions

### Implications

The interaction between fertilizer and ground cover demonstrates that fertilizer application doesn't simply shift overall OSE abundance—it changes *where* along the ground cover gradient OSE reach peak density. This has important implications for management strategies, as optimal ground cover targets may differ between fertilized and unfertilized fields.

Simpler models (null, linear interaction, or single smooth) were poorly supported, justifying the inclusion of fertilizer-specific smooths in reporting and interpretation.

# Analysis 4: Deep dive into ground cover x fertilizer impact on OSE density

Test the effect of fertilizer on OSE_count, and whether ground cover (as a smooth, nonlinear function) mediates the fertilizer effect, while controlling for repeated measures at the farmer level and nuisance effect of region.

## Unified model

$$
\begin{align*}
\text{OSE\_count}_i &= \beta_0 \\
    &\quad + \beta_1\, \text{fertilizer\_treatment}_i \\
    &\quad + \beta_2\, \text{mission\_number}_i \\
    &\quad + \beta_3\, (\text{fertilizer\_treatment}_i \times \text{mission\_number}_i) \\
    &\quad + \beta_4\, \text{region}_i \\
    &\quad + f_{\text{fertilizer\_treatment}_i}(\text{percent\_ground\_cover}_i) \\
    &\quad + b_{\text{farmer}_i} \\
    &\quad + \varepsilon_i
\end{align*}
$$

-   fertilizer_treatment \* mission_number:
    -   Captures difference before and after fertilizer application.
    -   Models differential effect of fertilizer after application (mission_number 2/3 vs 1).
-   region:
    -   Included as a fixed variable.
-   s(percent_ground_cover, by=fertilizer_treatment):
    -   Factor-smooth allows ground cover x OSE_count relationship to vary by treatment.
    -   Tests the “mediation" hypothesis.
-   (1\|farmer):
    -   Random intercept for farmers, controls for repeated measures and paired plot structure.

```{r}

data <- data |>
  mutate(
    farmer = factor(farmer),
    mission_number = factor(mission_number),
    fertilizer_treatment = factor(fertilizer_treatment),
    region = factor(region)
  )

mod <- gam(
  ose_count ~ 
    fertilizer_treatment * mission_number + 
    region + 
    s(percent_ground_cover, by=fertilizer_treatment) + 
    s(farmer,bs='re'),
  data = data,
  family = tw(),
  select = TRUE
)

dev_exp <- summary(mod)$dev.expl * 100
```

```{r}

# Parametric terms (fixed effects)
tidy_mod <- broom::tidy(mod, parametric = TRUE)
kable(tidy_mod, caption = "Parametric (Fixed) Effects from GAM Model")

# Smooth terms
smooth_mod <- broom::tidy(mod, parametric = FALSE)
kable(smooth_mod, caption = "Smooth Terms from GAM Model")



```

This unified model explained `{r} round(dev_exp, 3)`% of the variation and was pretty well fit.

### ground cover x fertilizer impact on OSE density

```{r}
# Extract smooth estimates for each fertilizer treatment
ests <- gratia::smooth_estimates(mod, 
                                 select = 'percent', 
                                 partial = TRUE)

# Adjust to response scale
ests$adj_est <- ests$.estimate + coef(mod)[1]

# Plot treatment-specific relationships
ests |>
  ggplot(aes(x = percent_ground_cover)) +
  geom_ribbon(
    aes(ymin = adj_est - .se, ymax = adj_est + .se, 
        fill = fertilizer_treatment),
    alpha = 0.2
  ) +
  geom_line(
    aes(y = adj_est, color = fertilizer_treatment),
    linewidth = 1.2
  ) +
  labs(
    x = "Ground Cover (%)",
    y = "Estimated OSE Count",
    title = "Ground Cover Effects by Fertilizer Treatment",
    color = "Treatment",
    fill = "Treatment"
  ) +
  scale_color_met_d(name = "Demuth") +
  scale_fill_met_d(name = "Demuth") +
  theme_minimal()

```

### OSE density by treatment and mission number

```{r treatment_mission_plot}
treat_mission <- emmeans(mod,~'fertilizer_treatment * mission_number',type='response') |>
  as_tibble()



ggplot(treat_mission, 
       aes(x=mission_number, y=response, 
           color=fertilizer_treatment)) +
  geom_point(position=position_dodge(width=0.3),size=3) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0,
                position=position_dodge(width=0.3)) +
  scale_color_manual(values = c("control" = "black", "fertilized" = "darkgreen")) +
  labs(x="Mission Number", y="Estimated OSE Density", color="Fertilizer Treatment") +
  theme_minimal()




```

### OSE density by region

```{r unified_region_mod_plot}

region <- emmeans(mod,~'region',type='response') |>
  as_tibble()



ggplot(region, 
       aes(x=region, y=response)) +
  geom_point(position=position_dodge(width=0.3), size=3) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0,
                position=position_dodge(width=0.3)) +
  labs(x="Region", y="Estimated OSE Density") +
  theme_minimal()




```

## Sensitivity Model (Stratified by ground cover)

W build two models for \< 50% and \>= 50% ground cover.

```{r}

below_dat <- data |>
  filter(percent_ground_cover < 50)

above_dat <- data |>
  filter(percent_ground_cover >= 50)

below_mod <- gam(ose_count ~ 
    fertilizer_treatment * mission_number + 
    region + 
    s(farmer,bs='re'),
    family=tw(),
    data = below_dat)

above_mod <- gam(ose_count ~ 
    fertilizer_treatment * mission_number + 
    region + 
    s(farmer,bs='re'),
    family=tw(),
    data=above_dat)

```

Quick model summary information:

```{r}



# Parametric terms (fixed effects)
below_mod_tidy <- broom::tidy(below_mod, parametric = TRUE)
kable(below_mod_tidy, caption = "Fixed effects for < 50% ground cover GAM Model")

# Smooth terms
above_mod_tidy <- broom::tidy(above_mod, parametric = TRUE)
kable(above_mod_tidy, caption = "Fixed effects for > 50% ground cover GAM Model")




```

Check the diagnostic plots:

```{r}

gratia::appraise(below_mod)
gratia::appraise(above_mod)

```

Diagnostic plots look good enough. The below model fits poorer than the above mod, but thats likely because of the small sample size.

```{r}

library(emmeans)

# Extract EMMs for below 50%
below_emm <- emmeans(below_mod, ~ fertilizer_treatment * mission_number, type = "response")
below_emm <- as.data.frame(below_emm)
below_emm$split <- "<50% ground cover"

# Extract EMMs for >=50%
above_emm <- emmeans(above_mod, ~ fertilizer_treatment * mission_number, type = "response")
above_emm <- as.data.frame(above_emm)
above_emm$split <- "≥50% ground cover"

# Combine
emm_plot_data <- rbind(below_emm, above_emm)

emm_plot_data$fertilizer_treatment <- factor(emm_plot_data$fertilizer_treatment, 
                                             levels=c("control", "fertilized"))
emm_plot_data$split <- factor(emm_plot_data$split, 
                             levels=c("<50% ground cover", "≥50% ground cover"))

ggplot(emm_plot_data, 
       aes(x=mission_number, y=response, 
           color=fertilizer_treatment, 
           shape=split, 
           group=interaction(fertilizer_treatment, split))) +
  geom_point(position=position_dodge(width=0.3),size=5) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0,
                position=position_dodge(width=0.3)) +
  scale_color_manual(values = c("control" = "black", "fertilized" = "darkgreen")) +
  labs(x="Mission Number", y="Estimated OSE Density", color="Fertilizer Treatment", 
       shape="Ground Cover", linetype="Ground Cover") +
  theme_minimal()


```

Mission 3 in the \<50% ground cover has very wide CI bars likely do to low sample size. Plots with \< 50% ground cover tend to have a lower OSE count than plots with \>= 50% ground cover. With that said, the trend between missions and fertilizer treatments reflect what we see in the fertilizer_treatment x region plot in the unified model. I would propose given this trend, we should just stick with the unified model.