---
title: "2021 Locust Density Analysis"
subtitle: "Density patterns by treatment and region"
format: html
toc: true
---

## Summary

The goal of this page is to analyze locust density patterns across fertilizer treatments and regions. This analysis replicates key figures from the manuscript to ensure consistent theming and validates our data management approach. If you see errors, please let me know!

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("locust_density_treatment_region")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data
data <- load_ose_data(cache_processed = TRUE)
```

# Locust density and region by treatment

To assess how locust count changes between treatment (fertilized and control) and region (four regions), I built a generalized linear mixed model (family: Poisson) per mission as follows:

$$
\begin{align*}
\text{ose count}_{ij} =\ 
  &\beta_0 \\
  &+ \beta_1 \cdot \text{fertilizer treatment}_i \\
  &+ \beta_2 \cdot \text{region}_i \\
  &+ \beta_3 \cdot (\text{fertilizer treatment}_i \times \text{region}_i) \\
  &+ u_j \\
  &+ \varepsilon_{ij}
\end{align*}
$$

Where: - $\text{ose count}_{ij}$: observed ose count for observation $i$ from farmer $j$ - $\text{fertilizer treatment}_i$: fertilizer treatment group of observation $i$ - $\text{region}_i$: region of observation $i$ - $(\text{fertilizer treatment}_i \times \text{region}_i)$: interaction between fertilizer treatment and region - $u_j$: random effect for farmer $j$, $u_j \sim N(0, \sigma^2_{farmer})$ - $\varepsilon_{ij}$: residual error

Below are the diagnostic plots for these models:

```{r computing-ose-density-models}
#| echo: false
#| message: false
#| warning: false

# Split data into a list of data based on mission date
model_data <- data |>
    group_split(mission_number)

# Create lists to store objects
mission_mods <- list()
mission_diagnostic_plots <- list()
mission_emmeans <- list()

# Loop through missions: create model, diagnostic plot, and emmeans with caching
for (i in seq_along(model_data)){
    diagnostic_plot_title <- paste0('Mission ', i)
    
    # Cache each mission model separately
    mission_mods[[i]] <- cache_model(
        model_name = paste0("locust_density_mission_", i),
        data = model_data[[i]],
        fit_function = create_glmm_model
    )
    
    mission_diagnostic_plots[[i]] <- plot_model_diagnostics(mission_mods[[i]], plot_title = diagnostic_plot_title)
    mission_emmeans[[i]] <- get_emmeans_and_pairs(mission_mods[[i]], mission_num = i)
}

names(mission_mods) <- c('m1', 'm2', 'm3')
names(mission_diagnostic_plots) <- c('m1', 'm2', 'm3')
names(mission_emmeans) <- c('m1', 'm2', 'm3')
```

```{r plot-diagnostic-plots}
#| echo: false
#| message: false
#| warning: false

diagnostic_plots <- (mission_diagnostic_plots$m1 +
    mission_diagnostic_plots$m2) /
    (mission_diagnostic_plots$m3 +
    patchwork::plot_spacer())

# Display diagnostic plots
diagnostic_plots

# Save model summaries for each mission (tables only)
for (i in 1:3) {
  save_model_summary(mission_mods[[i]], paste0("mission_", i, "_model_summary"))
  save_emmeans_table(mission_emmeans[[i]]$emmeans, paste0("mission_", i, "_emmeans"))
  save_data_table(mission_emmeans[[i]]$pairwise, paste0("mission_", i, "_pairwise_comparisons"))
}
```

These models are not perfect, but I think they are adequate for this analysis. Happy to investigate further if you would like me to!

## OSE density by treatment and region (option 1)

```{r density-option-1-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 5       
#| fig-height: 10      
#| fig-retina: 2      

# Combine model results with emmeans
mission_emms <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    df <- mission_emmeans[[nm]]$emmeans
    df$source <- nm
    df
  }),
  .id = 'mission_numer' 
)

count_plot_option_1 <- plot_mission_density(
    data |> drop_na(ose_count, fertilizer_treatment),
    mission_emms,
    ncol = 1,
    regions = ALT_STUDY_REGIONS,
    legend_position = 'bottom'
) 

# Display the plot
count_plot_option_1

# Save the combined emmeans data (table only)
save_data_table(mission_emms, "combined_mission_emmeans")
```

## OSE density by treatment and region (option 2)

Instead of a plot per mission, let's plot the missions together (x-axis) and see how region counts shifted through time

```{r count-plot-option-2}
#| echo: false
#| message: false
#| warning: false

count_plot_option_2 <- plot_emmeans(mission_emms,
    color_palette = DEFAULT_PALETTE,
    ylab = "OSE count",
    xlab = "Mission")

count_plot_option_2
```

## OSE density by teatment and region model summaries (tabulated)

Here are the model summaries and pairwise comparisons. This will be *across* missions, but please keep in mind each mission had it's own model. So we cant make across-mission comparions just yet.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
library(broom.mixed)
mission_mod_summary <- bind_rows(
  lapply(names(mission_mods), function(nm) {
    # Add the list element name as a column:
    df <- mission_mods[[nm]] |> tidy()
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

mission_emms <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    # Add the list element name as a column:
    df <- mission_emmeans[[nm]]$emmeans
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

mission_pairs <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    # Add the list element name as a column:
    df <- mission_emmeans[[nm]]$pairwise
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

library(gt)
library(dplyr)


mission_mod_summary <- mission_mod_summary |>
  arrange(mission_number) |>
  mutate(effect = ifelse(effect == 'ran_pars', 'random', 'fixed'),
    term = ifelse(term == 'sd__(Intercept)','farmer',term),
    term = gsub('fertilizer_treatment|region','',term)) |>
  select(!c(component,group))



mission_mod_summary |>
  select(!source) |>
  gt(groupname_col = "mission_number") |>
  fmt_number(
    columns = c(estimate, std.error, statistic, p.value),
    decimals = 3
  ) |>
  cols_label(
    mission_number = "Mission",
    effect = "Effect",
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. Error",
    statistic = "Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "OSE density by treatment and region by mission"
  )

```

Here are the pairwise comparisons between region and treatment by mission number. Fair warning, its a large table since there are a lot of regions.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
mission_pairs |>
  arrange(contrast) |>
  select(!source) |>
  gt(groupname_col = "mission_number") |>
  fmt_number(
    columns = c(ratio, SE, z.ratio, p.value),
    decimals = 3
  ) |>
  tab_header(
    title = "OSE density model pairwise comparisons by mission"
  ) 

```

# Table: annual OSE density by region and field type

The data discrepancies seen the the OSE count x ground cover are really shown in this table as well. We can see that the trends are the same (e.g. fertilized plots have less than control), but the actual averages are totally different.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
ose_dat <- data |>
  select(region, fertilizer_treatment, ose_count) |>
  mutate(region = factor(region, levels = c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')),
         fertilizer_treatment = factor(fertilizer_treatment, levels = c('fertilized', 'control'))) |>
  drop_na(ose_count, fertilizer_treatment) |>
  group_by(region, fertilizer_treatment) |>
  summarize(ose_mean = mean(ose_count, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = region, values_from = ose_mean)

# Create the gt table
ose_dat |>
  gt(rowname_col = "fertilizer_treatment") |>
  tab_header(
    title = "Mean OSE Count by Region and Fertilizer Treatment"
  ) |>
  fmt_number(
    columns = everything(),
    decimals = 2
  ) |>
  cols_label(
    fertilizer_treatment = "Treatment",
    `Saint Louis` = "Saint Louis",
    Thies = "Thies",
    Fatick = "Fatick",
    Kaffrine = "Kaffrine"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )


```