---
title: "2021 Locust × Ground Cover"
format: html
toc: true
---

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("locust_density_ground_cover")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data
data <- load_ose_data(cache_processed = TRUE)
```

We would like to see how density is influenced by ground cover. To do this I am building a generalized additive model (family: tweedie). Following this form:

The generalized additive model (GAM) fitted is:

$$
\begin{align*}
\text{ose\_count}_i = \\
  &\alpha + \\ 
  &+ f(\text{percent\_ground\_cover}_i,\, \text{region\_treat}_i) +  \\
  &+ b_{\text{farmer}_i} + \\
  &+ \varepsilon_i
\end{align*}
$$

where:

-   $\alpha$ is the intercept,
-   $f(\text{percent\_ground\_cover},\, \text{region\_treat})$ is a smooth function (tensor product, using factor-smooth interaction) over percent ground cover and region treatment,
-   $b_{\text{farmer}_i}$ is a random effect for each farmer,
-   $\varepsilon_i$ is the residual error.

```{r constructing-gam-model}
#| echo: false
#| message: false
#| warning: false

ground_cover_dat <- data |>
    dplyr::select(region, farmer, fertilizer_treatment, mission_number, ose_count, percent_ground_cover) |>
    mutate(
        region_treat = factor(paste0(region, '_', fertilizer_treatment)),
        farmer = factor(farmer)
    ) |>
    drop_na(ose_count, fertilizer_treatment)

# Use model caching for the GAM (can take several minutes to fit)
ground_cover_mod <- cache_model(
    model_name = "locust_ground_cover_gam",
    data = ground_cover_dat,
    fit_function = function(data, formula, nthreads, family) {
        build_gam(data, 
                 formula = formula, 
                 nthreads = nthreads, 
                 family = family)
    },
    # Simpler model for CI environments (faster fitting)
    fit_function_simple = function(data, formula, nthreads, family) {
        # Use simpler GAM with reduced complexity for CI
        build_gam(data,
                 formula = ose_count ~ te(percent_ground_cover, region_treat, bs = 'fs', k = 5) + s(farmer, bs = 're') +
                 mission_number,
                 nthreads = 1,
                 family = family)
    },
    formula = ose_count ~ te(percent_ground_cover, region_treat, bs = 'fs', k = 10) + s(farmer, bs = 're'),
    nthreads = 4,  # Reduced from 9 for better compatibility
    family = mgcv::tw()
)
```

The model diagnostics look pretty good! I tried to build a poisson model, but the data was overdispersed. Instead I went with a tweedie model which does a pretty good job at handling this type of data:

```{r gam model diagnostics}
#| echo: false
#| message: false
#| warning: false


appraise(ground_cover_mod)
```

However, I am not able to reproduce the smooths seen in the manuscript sent to me. This is likely due to data differences as I am not sure what was included in the initial modeling approach. **Roughly** we can see similiar trends in Kaffrine, Fatick, and Saint Louis. But Thies definitely has a different trend. Additionally, there seems to be different trends between control and fertilized treatments as comapred to the current plots in the manuscript. Not sure what to think here.

```{r plotting gam model results}
#| echo: false
#| message: false
#| warning: false

gam_plot <- plot_gam_smooths_gratia(ground_cover_mod)

gam_plot
```

## GAM model summary

```{r}
#| echo: false
#| message: false
#| warning: false


broom.mixed::tidy(ground_cover_mod) |>
  gt() |>
  fmt_number(
    columns = c(edf, ref.df, statistic, p.value),
    decimals = 2
  ) |>
  cols_label(
    term = "Term",
    edf = "Estimated DF",
    ref.df = "Reference DF",
    statistic = "F Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "GAM model Summary: OSE Density x ground Cover"
  ) |>
  tab_options(
    table.font.size = "small",
    row.striping.background_color = "#F6F6EF"
  ) |>
  as_raw_html()


```

# New analysis (12/26/2025)

I built a GAM with a Tweedie family and log link to model `ose_count`. The model includes a smooth for percent ground cover and random effects for region, fertilizer treatment, mission number, and farmer.

The relationship between ground cover and OSE count is non-linear, peaking at around 50% ground cover—OSE numbers are lower at both low and high cover. Region, fertilizer, mission, and farmer all accounted for substantial variation.

Overall, the model explains about 52% of the deviance in the data.

```{r running_new_model}


data <- data |>
    select(year,region,farmer,farmer_gender,fertilizer_treatment,mission_number,ose_count,percent_ground_cover) |>
    mutate(fertilizer_treatment = factor(fertilizer_treatment),
           region = factor(region),
           farmer = factor(farmer),
           mission_number = factor(mission_number),
           region_treat = factor(paste0(region, '_', fertilizer_treatment))
           )

data |> 
  glimpse()


mod <- bam(
  ose_count ~ 
    s(percent_ground_cover) + 
    s(region, bs="re") +
    s(fertilizer_treatment, bs="re") +
    s(mission_number, bs="re") +
    s(farmer, bs="re"),
  data = data,
  family = tw()
)

gratia::appraise(mod)

```

Here is the final modeled results for the relationship between ose_count and ground cover

```{r plotting_estimates}



est <- smooth_estimates(mod, select = "percent_ground_cover",partial_match=TRUE)

est$adj_est <- est$.estimate + coef(mod)[1]


ggplot(est, aes(x = percent_ground_cover, y = adj_est)) +
  geom_ribbon(aes(ymin = adj_est - .se, ymax = adj_est + .se), alpha = 0.2) +
  geom_line() +
  labs(
    x = "Percent Ground Cover",
    y = "Estimated OSE Count"
  ) +
  theme_minimal()



```

## ground cover by mission

in general, ground cover increased through the year. It does not really seem theres a big difference between regions.

```{r}

data |>
  ggplot(aes(x=mission_number,y=percent_ground_cover)) +
  geom_boxplot(outlier.size=-1) +
  geom_jitter(width=0.2,height=0) +
  facet_wrap(~region) +
  ggtitle('ground cover by region') +
  xlab('mission number') +
  ylab('ground cover (%)')


```



## ground cover by mission and fertilizer treatment


```{r ground_cover_fertilizer}

data |>
  drop_na(percent_ground_cover,fertilizer_treatment) |>
  ggplot(aes(x=mission_number, y=percent_ground_cover, fill=fertilizer_treatment)) +
  geom_boxplot(
    position = position_dodge(width = 0.8),
    outlier.shape = NA
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.4,jitter.height=0),
    alpha = 0.7,
    size = 1
  ) +
  facet_wrap(~region) +
  scale_color_brewer(palette = "Set2") +
  ggtitle('Ground cover by region') +
  xlab('Mission number') +
  ylab('Ground cover (%)') +
  theme_minimal()


mod <- glmmTMB(percent_ground_cover~fertilizer_treatment*mission_number,data=data)
summary(mod)
```

Ground cover wasnt different between fertilizer treatment, but it was between missions (makes sense). There was no interactive effect between fertilizer treatment and mission.


## OSE count by ground cover and fertilizer treatment


So agnostic to region, there are

```{r}

mod <- glmmTMB(ose_count ~ percent_ground_cover + fertilizer_treatment,data=data)

null_mod <- bam(
  ose_count ~ 
    1 +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data,
  family = tw(),
  select=TRUE
)

factor_smooth_mod <- bam(
  ose_count ~ 
    te(percent_ground_cover, by = fertilizer_treatment,k=30) +
    fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data,
  family = tw(),
  select=TRUE
)

linear_mod <- bam(
  ose_count ~ 
    percent_ground_cover *
    fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data,
  family = tw(),
  select=TRUE
)

without_fertilizer_mod <- bam(
  ose_count ~ 
    te(percent_ground_cover,k=30) +
    fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data,
  family = tw(),
  select=TRUE
)


aic_tbl <- AIC(
  null_mod,
  factor_smooth_mod,
  without_fertilizer_mod,
  linear_mod
) |>
  as_tibble(rownames = "model") |>
  mutate(
    model_name = case_when(
      model == "null_mod" ~ "Null Model",
      model == "factor_smooth_mod" ~ "Smooth by Fertilizer Treatment",
      model == "without_fertilizer_mod" ~ "No Fertilizer Treatment Smooth",
      model == "linear_mod" ~ "Linear Interaction",
      TRUE ~ model
    )
  ) |>
  mutate(delta_AIC = AIC - min(AIC)) |>
  arrange(delta_AIC) |>
  select(model_name, AIC, delta_AIC)

aic_tbl

```



So the model with a smooth term for both fertilizer treatments was selected by AIC. This means that fertilizer treatment has an impact on the relationship between ose_count and ground_cover. lets plot out this relationship.


```{r}

ests <- gratia::smooth_estimates(factor_smooth_mod,select='percent',partial=TRUE)


ests$adj_est <- ests$.estimate + coef(factor_smooth_mod)[1] #This is how I adjust to TWS units

ests %>% # not sigificant in the model
  ggplot(aes(x=percent_ground_cover)) +
  geom_ribbon(aes(ymin=adj_est-.se, ymax=adj_est+.se,fill=fertilizer_treatment),alpha=.2) +
  geom_line(aes(y = adj_est,color=fertilizer_treatment),size=1) + theme_pubr() +
  xlab("ground cover (%)") +
  ylab('OSE count') +
  scale_color_met_d(name="Demuth") +
  scale_fill_met_d(name="Demuth")


```



Of all the models tested, the one allowing the effect of ground cover to vary by fertilizer treatment (“smooth by fertilizer treatment”) was the best based on AIC. This means the shape of the relationship between ground cover and OSE count depends on whether the field was fertilized or not.

OSE counts in fertilized plots peak at around 40% ground cover, whereas in controls, the peak shifts to about 60–70%. That suggests fertilizer treatment modifies how OSE respond to ground cover levels perhaps through effects on crop vigor, species composition, or microclimate.

Model selection supports including an interaction between fertilizer and ground cover, highlighting that fertilizer status doesn’t just change overall abundance but alters where OSE are most dense along the ground cover gradient. Simpler models (single smooth, linear interaction, or null) were much less supported. These results justify showing the separate smooths by fertilizer in the main text.