---
title: "2021 Locust Damage Analysis"
subtitle: "Density patterns by treatment and region"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

## Summary

The goal of this page is to analyze locust damage patterns across fertilizer treatments and regions. This analysis replicates key figures from the manuscript to ensure consistent theming and validates our data management approach. If you see errors, please let me know!

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
# update to refelct damage
set_current_analysis("locust_damage_treatment_region")

# Load gt and webshot2 for table export
library(gt)
library(webshot2)

# Helper function to save gt tables as PNG and CSV
save_gt_png <- function(gt_obj, filename, subfolder = "locust_damage_treatment_region") {
  # Create output directories
  png_dir <- here::here("outputs", "tables", subfolder, "png")
  csv_dir <- here::here("outputs", "tables", subfolder, "csv")
  dir.create(png_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Save PNG
  png_path <- file.path(png_dir, filename)
  gtsave(gt_obj, png_path, vwidth = 3000)
  
  # Save CSV - extract the underlying data from the gt object
  csv_filename <- sub("\\.png$", ".csv", filename)
  csv_path <- file.path(csv_dir, csv_filename)
  
  # Extract data from gt object
  gt_data <- gt_obj$`_data`
  write_csv(gt_data, csv_path)
  
  return(invisible(list(png = png_path, csv = csv_path)))
}
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data
data <- load_ose_data(cache_processed = TRUE)

data <- data |>
    mutate(ose_damage_percent = ose_damage_percent/100)
```

# Locust density and region by treatment

To assess how locust damage percent changes between treatment (fertilized and control) and region (four regions), I built a generalized linear mixed model (family: Poisson) per mission as follows:

*add formula here*

Below are the diagnostic plots for these models:

```{r computing-ose-density-models}
#| echo: false
#| message: false
#| warning: false

# Split data into a list of data based on mission date
model_data <- data |>
    group_split(mission_number)

# Create lists to store objects
mission_mods <- list()
mission_diagnostic_plots <- list()
mission_emmeans <- list()

# Loop through missions: create model, diagnostic plot, and emmeans with caching
for (i in seq_along(model_data)){
    diagnostic_plot_title <- paste0('Mission ', i)
    
    # Cache each mission model separately
    mission_mods[[i]] <- cache_model(
        model_name = paste0("locust_damage_mission_", i),
        data = model_data[[i]],
        fit_function = create_damage_glmm_model
    )
    
    mission_diagnostic_plots[[i]] <- plot_model_diagnostics(mission_mods[[i]], plot_title = diagnostic_plot_title)
    mission_emmeans[[i]] <- get_emmeans_and_pairs(mission_mods[[i]], mission_num = i)
}

names(mission_mods) <- c('m1', 'm2', 'm3')
names(mission_diagnostic_plots) <- c('m1', 'm2', 'm3')
names(mission_emmeans) <- c('m1', 'm2', 'm3')
```

```{r plot-diagnostic-plots}
#| echo: false
#| message: false
#| warning: false

diagnostic_plots <- (mission_diagnostic_plots$m1 +
    mission_diagnostic_plots$m2) /
    (mission_diagnostic_plots$m3 +
    patchwork::plot_spacer())

# Display diagnostic plots
diagnostic_plots

# Save model summaries for each mission (tables only)
for (i in 1:3) {
  save_model_summary(mission_mods[[i]], paste0("mission_", i, "_model_summary"))
  save_emmeans_table(mission_emmeans[[i]]$emmeans, paste0("mission_", i, "_emmeans"))
  save_data_table(mission_emmeans[[i]]$pairwise, paste0("mission_", i, "_pairwise_comparisons"))
}
```

The tweedie models look like they capture the overdispersion quite well. There is a lot of zeros in this data.

## OSE damage by treatment and region

```{r density-option-1-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 5       
#| fig-height: 10      
#| fig-retina: 2      

# Combine model results with emmeans
mission_emms <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    df <- mission_emmeans[[nm]]$emmeans
    df$source <- nm
    df
  }),
  .id = 'mission_numer' 
)

count_plot_option_1 <- plot_mission_damage(
    data |> drop_na(ose_damage_percent, fertilizer_treatment),
    mission_emms,
    ncol = 1,
    regions = c("Kaffrine", "Thies", "Fatick", "Saint Louis"),
    legend_position = 'bottom'
) 

# Display the plot
count_plot_option_1

# Save the combined emmeans data (table only)
save_data_table(mission_emms, "combined_mission_emmeans")
```

## OSE damage by teatment and region model summaries (tabulated)

Here are the model summaries and pairwise comparisons. This will be *across* missions, but please keep in mind each mission had it's own model. So we cant make across-mission comparions just yet.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
library(broom.mixed)
mission_mod_summary <- bind_rows(
  lapply(names(mission_mods), function(nm) {
    # Add the list element name as a column:
    df <- mission_mods[[nm]] |> tidy()
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

mission_emms <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    # Add the list element name as a column:
    df <- mission_emmeans[[nm]]$emmeans
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

mission_pairs <- bind_rows(
  lapply(names(mission_emmeans), function(nm) {
    # Add the list element name as a column:
    df <- mission_emmeans[[nm]]$pairwise
    df$source <- nm
    df
  }),
  .id = 'mission_number' 
)

library(gt)
library(dplyr)



mission_mod_summary <- mission_mod_summary |>
  arrange(mission_number) |>
  mutate(
    effect = ifelse(effect == "ran_pars", "random", "fixed"),
    term = ifelse(term == "sd__(Intercept)", "farmer", term),
    term = gsub("fertilizer_treatment|region", "", term)
  ) |>
  select(-any_of(c("component", "group")))


tbl_damage_models <- mission_mod_summary |>
  select(!source) |>
  gt(groupname_col = "mission_number") |>
  fmt_number(
    columns = c(estimate, std.error, statistic, p.value),
    decimals = 3
  ) |>
  cols_label(
    mission_number = "Mission",
    effect = "Effect",
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. Error",
    statistic = "Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "OSE density by treatment and region by mission"
  )

save_gt_png(tbl_damage_models, "ose_damage_model_summaries_by_mission.png")
tbl_damage_models

```

Here are the pairwise comparisons between region and treatment by mission number. Fair warning, its a large table since there are a lot of regions.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
tbl_damage_pairwise <- mission_pairs |>
  arrange(contrast) |>
  select(!source) |>
  gt(groupname_col = "mission_number") |>
  fmt_number(
    columns = c(ratio, SE, z.ratio, p.value),
    decimals = 3
  ) |>
  tab_header(
    title = "OSE density model pairwise comparisons by mission"
  )

save_gt_png(tbl_damage_pairwise, "ose_damage_pairwise_comparisons_by_mission.png")
tbl_damage_pairwise 

```

# Table: annual OSE damage by region and field type

The data discrepancies seen the the OSE count x ground cover are really shown in this table as well. We can see that the trends are the same (e.g. fertilized plots have less than control), but the actual averages are totally different.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
ose_dat <- data |>
  select(region, fertilizer_treatment, ose_damage_percent) |>
  mutate(region = factor(region, levels = c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')),
         fertilizer_treatment = factor(fertilizer_treatment, levels = c('fertilized', 'control'))) |>
  drop_na(ose_damage_percent, fertilizer_treatment) |>
  group_by(region, fertilizer_treatment) |>
  summarize(ose_mean = mean(ose_damage_percent, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = region, values_from = ose_mean)

# Create the gt table
tbl_damage_summary <- ose_dat |>
  gt(rowname_col = "fertilizer_treatment") |>
  tab_header(
    title = "Mean OSE damage by Region and Fertilizer Treatment"
  ) |>
  fmt_number(
    columns = everything(),
    decimals = 4
  ) |>
  cols_label(
    fertilizer_treatment = "Treatment",
    `Saint Louis` = "Saint Louis",
    Thies = "Thies",
    Fatick = "Fatick",
    Kaffrine = "Kaffrine"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )

save_gt_png(tbl_damage_summary, "mean_ose_damage_by_region_treatment.png")
tbl_damage_summary


```