---
title: "2021 Locust × Ground Cover"
format: html
toc: true
---

::: callout-note
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Core packages for visualization
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(ggpubr)
  library(mgcv)
  library(gratia)
  library(emmeans)
  library(DHARMa)
  library(broom)
  library(broom.mixed)
  library(gt)
  library(webshot2)
  library(MetBrewer)
  library(glmmTMB)
  library(knitr)
})

# Plot defaults
theme_set(theme_pubr())
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.retina = 2
)

# Source project functions
function_files <- list.files(here("R", "functions"), pattern = "\\.R$", full.names = TRUE)
for (f in function_files) source(f, local = TRUE)

# Project configuration constants (from _targets.R)
STUDY_REGIONS <- c('Saint-Louis', 'Thiès', 'Fatick', 'Kaffrine')
ALT_STUDY_REGIONS <- c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')
MISSION_LABELS <- c('Mission 1 (July)', 'Mission 2 (September)', 'Mission 3 (October)')
DEFAULT_PALETTE <- "Degas"
FERTILIZER_COLORS <- c('control' = 'black', 'fertilized' = 'dark green')
DEFAULT_POINT_SIZE <- 5
DEFAULT_EMMEAN_POINT_SIZE <- 8
DEFAULT_FIGURE_DPI <- 2

# Load gt and webshot2 for table export
library(gt)
library(webshot2)

# Helper function to save gt tables as PNG and CSV
save_gt_png <- function(gt_obj, filename, subfolder = "locust_density_ground_cover") {
  # Create output directories
  png_dir <- here::here("outputs", "tables", subfolder, "png")
  csv_dir <- here::here("outputs", "tables", subfolder, "csv")
  dir.create(png_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Save PNG
  png_path <- file.path(png_dir, filename)
  gtsave(gt_obj, png_path, vwidth = 3000)
  
  # Save CSV - extract the underlying data from the gt object
  csv_filename <- sub("\\.png$", ".csv", filename)
  csv_path <- file.path(csv_dir, csv_filename)
  
  # Extract data from gt object
  gt_data <- gt_obj$`_data`
  write_csv(gt_data, csv_path)
  
  return(invisible(list(png = png_path, csv = csv_path)))
}
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data from targets pipeline
data <- read_csv(here("data", "processed", "ose_data_processed.csv"), show_col_types = FALSE)

```

## Overview

This analysis explores how locust (OSE) density is influenced by ground cover across different regions and fertilizer treatments. We use generalized additive models (GAMs) with a Tweedie family distribution to handle the overdispersed count data.

------------------------------------------------------------------------

# Analysis 1: Region-Treatment Interaction Model (Initial Approach)

This initial model uses a tensor product smooth to examine how the relationship between ground cover and OSE count varies across region-treatment combinations.

## Model Specification

$$
\begin{align*}
\text{ose\_count}_i &= \alpha + f(\text{percent\_ground\_cover}_i, \text{region\_treat}_i) + b_{\text{farmer}_i} + \varepsilon_i
\end{align*}
$$

where:

-   $\alpha$ is the intercept,
-   $f(\text{percent\_ground\_cover},\, \text{region\_treat})$ is a smooth function (tensor product, using factor-smooth interaction) over percent ground cover and region treatment,
-   $b_{\text{farmer}_i}$ is a random effect for each farmer,
-   $\varepsilon_i$ is the residual error.

## Model Fitting

```{r region-treatment-model}
#| echo: false
#| message: false
#| warning: false

# Prepare data for region-treatment interaction model
ground_cover_dat <- data |>
    select(region, farmer, fertilizer_treatment, mission_number, 
           ose_count, percent_ground_cover) |>
    mutate(
        region_treat = factor(paste0(region, '_', fertilizer_treatment)),
        farmer = factor(farmer)
    ) |>
    drop_na(ose_count, fertilizer_treatment)

# Load ground cover GAM from targets pipeline
ground_cover_mod <- readRDS(here::here("data", "model_objects", "ground_cover_mediation.rds"))
```

The unified ground cover mediation model is a GAM with by-variable smooths that allows the relationship between ground cover and OSE density to vary by fertilizer treatment. Diagnostics look good:

```{r gam model diagnostics}
#| echo: false
#| message: false
#| warning: false

gratia::appraise(ground_cover_mod)
```

## Model Results

The unified model shows how ground cover mediates fertilizer effects on OSE density. The smoothed relationships show different patterns between control and fertilized treatments.

**Note:** These results may differ from earlier manuscript versions due to data processing differences.

```{r region-treatment-results}
#| echo: false
#| message: false
#| warning: false

# Extract smooth estimates for each fertilizer treatment using by-variable smooths
ests <- gratia::smooth_estimates(ground_cover_mod, 
                                 select = 'percent', 
                                 partial = TRUE)

# Adjust to response scale
ests$adj_est <- ests$.estimate + coef(ground_cover_mod)[1]

# Plot treatment-specific relationships
gam_plot <- ests |>
  ggplot(aes(x = percent_ground_cover)) +
  geom_ribbon(
    aes(ymin = adj_est - .se, ymax = adj_est + .se, 
        fill = fertilizer_treatment),
    alpha = 0.2
  ) +
  geom_line(
    aes(y = adj_est, color = fertilizer_treatment),
    linewidth = 1.2
  ) +
  labs(
    x = "Ground Cover (%)",
    y = "Estimated OSE Density",
    title = "Ground Cover Effects by Fertilizer Treatment",
    color = NULL,
    fill = NULL
  ) +
  scale_color_manual(values = c("control" = "black", "fertilized" = "dark green")) +
  scale_fill_manual(values = c("control" = "black", "fertilized" = "dark green")) +
  theme_pubr() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14, face = "bold")
  )

gam_plot

gam_plot
```

## Model Summary Statistics

```{r region-treatment-summary}
#| echo: false
#| message: false
#| warning: false

tbl_gc_model <- broom.mixed::tidy(ground_cover_mod) |>
  gt() |>
  fmt_number(
    columns = c(edf, ref.df, statistic, p.value),
    decimals = 2
  ) |>
  cols_label(
    term = "Term",
    edf = "Estimated DF",
    ref.df = "Reference DF",
    statistic = "F Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "GAM Model Summary: OSE Density × Ground Cover (Region-Treatment Model)"
  ) |>
  tab_options(
    table.font.size = "small",
    row.striping.background_color = "#F6F6EF"
  )

save_gt_png(tbl_gc_model, "gam_ose_density_ground_cover_model_summary.png")
tbl_gc_model |> as_raw_html()


```

------------------------------------------------------------------------

# Analysis 2: Additive Random Effects Model (Current Analysis)

*Updated: December 26, 2025*

This refined approach uses a simpler model structure with additive random effects rather than complex interactions. This model includes:

-   A smooth term for percent ground cover
-   Random effects for region, fertilizer treatment, mission number, and farmer

Key findings:

-   The relationship between ground cover and OSE count is non-linear, peaking around 50% ground cover
-   OSE counts are lower at both low and high ground cover
-   The model explains approximately 52% of deviance in the data

## Exploratory Analysis (Deprecated)

\`\`\`\`{=html} <!-- 
This section contains an earlier exploratory additive model that has been superseded 
by the unified mediation model below. Kept for reference but not rendered.

```{r additive-model-fit}
#| echo: false
#| message: false
#| warning: false
#| eval: false


# Prepare data with all factors
data_filtered <- data |>
    select(year, region, farmer, farmer_gender, fertilizer_treatment, 
           mission_number, ose_count, percent_ground_cover) |>
    mutate(
      fertilizer_treatment = factor(fertilizer_treatment),
      region = factor(region),
      farmer = factor(farmer),
      mission_number = factor(mission_number)
    )

# Fit additive model with random effects
mod <- bam(
  ose_count ~ 
    s(percent_ground_cover) + 
    s(region, bs = "re") +
    s(fertilizer_treatment, bs = "re") +
    s(mission_number, bs = "re") +
    s(farmer, bs = "re"),
  data = data_filtered,
  family = tw()
)

# Model diagnostics
gratia::appraise(mod)

```

```{r additive-model-ground-cover-effect}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| eval: false

# Extract smooth estimates for ground cover
est <- smooth_estimates(mod, select = "percent_ground_cover", partial_match = TRUE)

# Adjust estimates to response scale
est$adj_est <- est$.estimate + coef(mod)[1]

# Plot the relationship
ggplot(est, aes(x = percent_ground_cover, y = adj_est)) +
  geom_ribbon(aes(ymin = adj_est - .se, ymax = adj_est + .se), 
              alpha = 0.2, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    x = "Ground Cover (%)",
    y = "Estimated OSE Count",
    title = "Effect of Ground Cover on OSE Density"
  ) +
  theme_minimal()
```
-->

## Ground Cover Patterns

```{r prepare-data-filtered}
#| echo: false
#| message: false
#| warning: false

# Prepare data with all factors for descriptive plots
data_filtered <- data |>
    select(year, region, farmer, farmer_gender, fertilizer_treatment, 
           mission_number, ose_count, percent_ground_cover) |>
    mutate(
      fertilizer_treatment = factor(fertilizer_treatment),
      region = factor(region),
      farmer = factor(farmer),
      mission_number = factor(mission_number)
    )
```

### Ground Cover by Mission

Ground cover increased throughout the growing season across all regions, with minimal regional differences.

```{r ground-cover-mission}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

data_filtered |>
  ggplot(aes(x = mission_number, y = percent_ground_cover)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.4) +
  facet_wrap(~region) +
  labs(
    title = "Ground Cover Progression Through Growing Season",
    x = "Mission Number",
    y = "Ground Cover (%)"
  ) +
  theme_minimal()
```

### Ground Cover by Fertilizer Treatment

Ground cover did not differ significantly between fertilizer treatments, but did vary between missions. No significant interaction was detected.

```{r ground-cover-fertilizer-treatment}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

data_filtered |>
  drop_na(percent_ground_cover, fertilizer_treatment) |>
  ggplot(aes(x = mission_number, y = percent_ground_cover, 
             fill = fertilizer_treatment)) +
  geom_boxplot(
    position = position_dodge(width = 0.8),
    outlier.shape = NA
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.4, jitter.height = 0),
    alpha = 0.5,
    size = 0.8
  ) +
  facet_wrap(~region) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Ground Cover by Fertilizer Treatment and Mission",
    x = "Mission Number",
    y = "Ground Cover (%)",
    fill = "Treatment"
  ) +
  theme_minimal()

# Statistical test for fertilizer × mission interaction
ground_cover_test <- glmmTMB(
  percent_ground_cover ~ fertilizer_treatment * mission_number,
  data = data_filtered
)
summary(ground_cover_test)
```

------------------------------------------------------------------------

````{=html}
<!--
# Analysis 3: Fertilizer-Specific Ground Cover Effects

*Model Comparison Analysis*

EXPLORATORY SECTION - Commented out as these models are not used in the manuscript.
The unified model from targets is the canonical version.

To determine whether the effect of ground cover on OSE count varies by fertilizer treatment, we compare multiple model structures using AIC.

## Model Comparison

```{r fertilizer-interaction-models}
#| echo: false
#| message: false
#| warning: false
#| eval: false

# Null model (baseline)
null_mod <- bam(
  ose_count ~ 
    1 +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Model with fertilizer-specific smooth for ground cover
factor_smooth_mod <- bam(
  ose_count ~ 
    te(percent_ground_cover, by = fertilizer_treatment, k = 30) +
    fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Linear interaction model
linear_mod <- bam(
  ose_count ~ 
    percent_ground_cover * fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Single smooth without fertilizer interaction
without_fertilizer_mod <- bam(
  ose_count ~ 
    te(percent_ground_cover, k = 30) +
    fertilizer_treatment +
    s(farmer, bs = "re") +
    s(mission_number, bs = "re"),
  data = data_filtered,
  family = tw(),
  select = TRUE
)

# Compare models using AIC
aic_tbl <- AIC(
  null_mod,
  factor_smooth_mod,
  without_fertilizer_mod,
  linear_mod
) |>
  as_tibble(rownames = "model") |>
  mutate(
    model_name = case_when(
      model == "null_mod" ~ "Null Model",
      model == "factor_smooth_mod" ~ "Smooth by Fertilizer Treatment",
      model == "without_fertilizer_mod" ~ "No Fertilizer Treatment Smooth",
      model == "linear_mod" ~ "Linear Interaction",
      TRUE ~ model
    )
  ) |>
  mutate(delta_AIC = AIC - min(AIC)) |>
  arrange(delta_AIC) |>
  select(model_name, AIC, delta_AIC)

aic_tbl
```

## Model Selection Results

The model with fertilizer-specific smooths was strongly supported by AIC.

## Fertilizer-Specific Effects

```{r fertilizer-specific-smooths}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| eval: false

# Extract smooth estimates for each fertilizer treatment
ests <- gratia::smooth_estimates(factor_smooth_mod, 
                                 select = 'percent', 
                                 partial = TRUE)

# Adjust to response scale
ests$adj_est <- ests$.estimate + coef(factor_smooth_mod)[1]

# Plot treatment-specific relationships
ests |>
  ggplot(aes(x = percent_ground_cover)) +
  geom_ribbon(
    aes(ymin = adj_est - .se, ymax = adj_est + .se, 
        fill = fertilizer_treatment),
    alpha = 0.2
  ) +
  geom_line(
    aes(y = adj_est, color = fertilizer_treatment),
    linewidth = 1.2
  ) +
  labs(
    x = "Ground Cover (%)",
    y = "Estimated OSE Count",
    title = "Ground Cover Effects by Fertilizer Treatment",
    color = "Treatment",
    fill = "Treatment"
  ) +
  scale_color_met_d(name = "Demuth") +
  scale_fill_met_d(name = "Demuth") +
  theme_minimal()
```

## Interpretation

### Key Findings

1.  **Model Selection**: The fertilizer-specific smooth model was strongly preferred.
-->
````

------------------------------------------------------------------------

2.  **Peak Densities**:

    -   **Fertilized plots**: OSE counts peak at approximately 40% ground cover
    -   **Control plots**: OSE counts peak at approximately 60–70% ground cover

3.  **Mechanism**: Fertilizer treatment likely modifies OSE responses through:

    -   Changes in crop vigor and nutritional quality
    -   Altered plant species composition
    -   Modified microclimate conditions

### Implications

The interaction between fertilizer and ground cover demonstrates that fertilizer application doesn't simply shift overall OSE abundance—it changes *where* along the ground cover gradient OSE reach peak density. This has important implications for management strategies, as optimal ground cover targets may differ between fertilized and unfertilized fields.

Simpler models (null, linear interaction, or single smooth) were poorly supported, justifying the inclusion of fertilizer-specific smooths in reporting and interpretation.

# Analysis 4: Deep dive into ground cover x fertilizer impact on OSE density

Test the effect of fertilizer on OSE_count, and whether ground cover (as a smooth, nonlinear function) mediates the fertilizer effect, while controlling for repeated measures at the farmer level and nuisance effect of region.

## Unified model

$$
\begin{align*}
\text{OSE\_count}_i &= \beta_0 \\
    &\quad + \beta_1\, \text{fertilizer\_treatment}_i \\
    &\quad + \beta_2\, \text{mission\_number}_i \\
    &\quad + \beta_3\, (\text{fertilizer\_treatment}_i \times \text{mission\_number}_i) \\
    &\quad + \beta_4\, \text{region}_i \\
    &\quad + f_{\text{fertilizer\_treatment}_i}(\text{percent\_ground\_cover}_i) \\
    &\quad + b_{\text{farmer}_i} \\
    &\quad + \varepsilon_i
\end{align*}
$$

-   fertilizer_treatment \* mission_number:
    -   Captures difference before and after fertilizer application.
    -   Models differential effect of fertilizer after application (mission_number 2/3 vs 1).
-   region:
    -   Included as a fixed variable.
-   s(percent_ground_cover, by=fertilizer_treatment):
    -   Factor-smooth allows ground cover x OSE_count relationship to vary by treatment.
    -   Tests the “mediation" hypothesis.
-   (1\|farmer):
    -   Random intercept for farmers, controls for repeated measures and paired plot structure.

```{r}
#| echo: false
#| message: false
#| warning: false

# Load the unified model from targets pipeline (already loaded above as ground_cover_mod)
mod <- ground_cover_mod
dev_exp <- summary(mod)$dev.expl * 100
```

```{r}
#| echo: false
#| message: false
#| warning: false

# Parametric terms (fixed effects)
tidy_mod <- broom::tidy(mod, parametric = TRUE)
knitr::kable(tidy_mod, caption = "Parametric (Fixed) Effects from GAM Model")

# Smooth terms
smooth_mod <- broom::tidy(mod, parametric = FALSE)
knitr::kable(smooth_mod, caption = "Smooth Terms from GAM Model")
```

This unified model explained `{r} round(dev_exp, 3)`% of the variation and was pretty well fit.

### ground cover x fertilizer impact on OSE density

```{r}
# Extract smooth estimates for each fertilizer treatment
ests <- gratia::smooth_estimates(mod, 
                                 select = 'percent', 
                                 partial = TRUE)

# Adjust to response scale
ests$adj_est <- ests$.estimate + coef(mod)[1]

# Plot treatment-specific relationships
ests |>
  ggplot(aes(x = percent_ground_cover)) +
  geom_ribbon(
    aes(ymin = adj_est - .se, ymax = adj_est + .se, 
        fill = fertilizer_treatment),
    alpha = 0.2
  ) +
  geom_line(
    aes(y = adj_est, color = fertilizer_treatment),
    linewidth = 1.2
  ) +
  labs(
    x = "Ground Cover (%)",
    y = "Estimated OSE Count",
    title = "Ground Cover Effects by Fertilizer Treatment",
    color = "Treatment",
    fill = "Treatment"
  ) +
  scale_color_met_d(name = "Demuth") +
  scale_fill_met_d(name = "Demuth") +
  theme_minimal()

```

### OSE density by treatment and mission number

```{r treatment_mission_plot}
treat_mission <- emmeans(mod,~'fertilizer_treatment * mission_number',type='response') |>
  as_tibble()



ggplot(treat_mission, 
       aes(x=mission_number, y=response, 
           color=fertilizer_treatment)) +
  geom_point(position=position_dodge(width=0.3),size=3) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0,
                position=position_dodge(width=0.3)) +
  scale_color_manual(values = c("control" = "black", "fertilized" = "darkgreen")) +
  labs(x="Mission Number", y="Estimated OSE Density", color="Fertilizer Treatment") +
  theme_minimal()




```

### OSE density by region

```{r unified_region_mod_plot}

region <- emmeans(mod,~'region',type='response') |>
  as_tibble()



ggplot(region, 
       aes(x=region, y=response)) +
  geom_point(position=position_dodge(width=0.3), size=3) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0,
                position=position_dodge(width=0.3)) +
  labs(x="Region", y="Estimated OSE Density") +
  theme_minimal()
```

````{=html}
<!--
## Sensitivity Model (Stratified by ground cover)

EXPLORATORY SECTION - Commented out as this approach is not used in the manuscript.

We build two models for < 50% and >= 50% ground cover.

```{r}
#| eval: false

below_dat <- data |>
  filter(percent_ground_cover < 50)

above_dat <- data |>
  filter(percent_ground_cover >= 50)

below_mod <- gam(ose_count ~ 
    fertilizer_treatment * mission_number + 
    region + 
    s(farmer,bs='re'),
    family=tw(),
    data = below_dat)

above_mod <- gam(ose_count ~ 
    fertilizer_treatment * mission_number + 
    region + 
    s(farmer,bs='re'),
    family=tw(),
    data=above_dat)
```

Quick model summary information:

```{r}
#| eval: false

summary(below_mod)
summary(above_mod)
```

Below 50% ground cover:

```{r}
#| eval: false

below_emm <- emmeans(below_mod, ~ fertilizer_treatment * mission_number, type='response') |>
  as_tibble()

below_emm$split <- "<50% ground cover"
below_emm
```

Above 50% ground cover:

```{r}
#| eval: false

above_emm <- emmeans(above_mod, ~ fertilizer_treatment * mission_number, type='response') |>
  as_tibble()

above_emm$split <- "≥50% ground cover"
above_emm
```

Combined plot:

```{r}
#| eval: false

# Combine
emm_plot_data <- rbind(below_emm, above_emm)

emm_plot_data$fertilizer_treatment <- factor(emm_plot_data$fertilizer_treatment, 
                                             levels=c("control", "fertilized"))
emm_plot_data$split <- factor(emm_plot_data$split, 
                             levels=c("<50% ground cover", "≥50% ground cover"))

ggplot(emm_plot_data, 
       aes(x=mission_number, y=response, 
           color=fertilizer_treatment, 
           shape=split, 
           group=interaction(fertilizer_treatment, split))) +
  geom_point(position=position_dodge(width=0.3),size=5) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0,
                position=position_dodge(width=0.3)) +
  scale_color_manual(values = c("control" = "black", "fertilized" = "darkgreen")) +
  labs(x="Mission Number", y="Estimated OSE Density", color="Fertilizer Treatment", 
       shape="Ground Cover", linetype="Ground Cover") +
  theme_minimal()
```

Mission 3 in the <50% ground cover has very wide CI bars likely due to low sample size.

```{r}
#| eval: false

# Parametric terms (fixed effects)
below_mod_tidy <- broom::tidy(below_mod, parametric = TRUE)
k nt r::kable(below_mod_tidy, caption = "Fixed effects for < 50% ground cover GAM Model")

# Smooth terms
above_mod_tidy <- broom::tidy(above_mod, parametric = TRUE)
k nt r::kable(above_mod_tidy, caption = "Fixed effects for > 50% ground cover GAM Model")
```

Check the diagnostic plots:

```{r}
#| eval: false

gratia::appraise(below_mod)
gratia::appraise(above_mod)
```

Diagnostic plots look good enough. The below model fits poorer than the above mod, but thats likely because of the small sample size.

```{r}
#| eval: false

library(emmeans)

# Extract EMMs for below 50%
below_emm <- emmeans(below_mod, ~ fertilizer_treatment * mission_number, type = "response")
below_emm <- as.data.frame(below_emm)
below_emm$split <- "<50% ground cover"

# Extract EMMs for >=50%
above_emm <- emmeans(above_mod, ~ fertilizer_treatment * mission_number, type = "response")
above_emm <- as.data.frame(above_emm)
above_emm$split <- "≥50% ground cover"

# Combine
emm_plot_data <- rbind(below_emm, above_emm)

emm_plot_data$fertilizer_treatment <- factor(emm_plot_data$fertilizer_treatment, 
                                             levels=c("control", "fertilized"))
emm_plot_data$split <- factor(emm_plot_data$split, 
                             levels=c("<50% ground cover", "≥50% ground cover"))

ggplot(emm_plot_data, 
       aes(x=mission_number, y=response, 
           color=fertilizer_treatment, 
           shape=split, 
           group=interaction(fertilizer_treatment, split))) +
  geom_point(position=position_dodge(width=0.3),size=5) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0,
                position=position_dodge(width=0.3)) +
  scale_color_manual(values = c("control" = "black", "fertilized" = "darkgreen")) +
  labs(x="Mission Number", y="Estimated OSE Density", color="Fertilizer Treatment", 
       shape="Ground Cover", linetype="Ground Cover") +
  theme_minimal()
```

Mission 3 in the <50% ground cover has very wide CI bars likely due to low sample size. Plots with < 50% ground cover tend to have a lower OSE count than plots with >= 50% ground cover. With that said, the trend between missions and fertilizer treatments reflect what we see in the fertilizer_treatment x region plot in the unified model. I would propose given this trend, we should just stick with the unified model.
-->
````

------------------------------------------------------------------------

# Summary

This exploratory analysis informed the development of the **unified ground cover mediation model** used in the manuscript. The unified model (loaded from targets pipeline above) provides the canonical analysis of how ground cover mediates fertilizer effects on grasshopper density.