---
title: "2021 Locust × Temperature"
format: html
toc: true
---

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("locust_density_temperature")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data and filter for control plots only
data <- load_ose_data(cache_processed = TRUE) |>
    filter(fertilizer_treatment == 'control') |>
    select(year, region, farmer, mission_number, ose_count, temperature)
```

# How is locust density related to temperature through the missions

We would like to model the influence of temperature on locust density. Let's plot the raw data by region to see how it falls. I also removed the fertilizer plots since this variable influences OSE counts.

```{r raw-data-viz}
#| echo: false
#| message: false
#| warning: false

raw_temp_plot <- data |>
    ggplot(aes(x = temperature, y = ose_count, color = region)) +
        geom_point() +
        geom_smooth(se = FALSE) +
        theme_pubr() +
        scale_color_met_d(name = 'Degas') +
        facet_wrap(~region, scales = 'free')

# Display the raw data plot
raw_temp_plot
```

The raw plots do not look very compelling, but let's construct a generalized additive model and see if there is a statistically significant trend.

```{r building-gam}
#| echo: false
#| message: false
#| warning: false

# Use model caching for the temperature GAM
mod <- cache_model(
    model_name = "locust_temperature_gam",
    data = data,
    fit_function = function(data) {
        mgcv::gam(
            ose_count ~
                s(temperature, region, bs = 'fs', k = 10) +
                s(farmer, bs = 're') +
                s(mission_number, bs = 're'),
            data = data,
            family = mgcv::tw(),
            select = TRUE
        )
    },
    # Simpler model for CI (reduced complexity)
    fit_function_simple = function(data) {
        mgcv::gam(
            ose_count ~
                s(temperature, region, bs = 'fs', k = 5) +
                s(farmer, bs = 're'),
            data = data,
            family = mgcv::tw(),
            select = TRUE
        )
    }
)
```

```{r}
#| echo: false
#| message: false
#| warning: false

temp_smooth_plot <- plot_temperature_smooth(mod,
    smooth = "s(temperature,region)",
    xvar = "temperature",
    group = "region",
    xlab = NULL,
    ylab = NULL,
    title = "locust density x temperature",
    palette = "Degas",
    ribbon_fill = "grey70",
    ribbon_alpha = 0.3) 

# Display and save the temperature smooth plot
temp_smooth_plot
save_temperature_smooth(mod,
    smooth = "s(temperature,region)",
    xvar = "temperature",
    group = "region",
    xlab = NULL,
    ylab = NULL,
    title = "locust density x temperature",
    palette = "Degas",
    ribbon_fill = "grey70",
    ribbon_alpha = 0.3,
    filename = "temperature_gam_smooths")

# Save model summary
save_model_summary(mod, "temperature_gam_summary")
```

There seems to be some positive realationship between temperature and locust denisty for Fatick, Saint Louis, and Thies. Kaffrine is a different relationship, however I wonder if this is partially still driven by the high densities in the first mission.

Here is the model summary for what it's worth:

```{r locust_temperature_model_summary}
#| echo: false
#| message: false
#| warning: false

model_summary_gt(mod, title = 'Locust x Temperature model summary', digits = 3) 
```