---
title: "2021 Locust × Temperature"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Core packages for visualization
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(ggpubr)
  library(mgcv)
  library(gratia)
  library(MetBrewer)
  library(gt)
})

# Plot defaults
theme_set(theme_pubr())
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.retina = 2
)

# Source project functions
function_files <- list.files(here("R", "functions"), pattern = "\\.R$", full.names = TRUE)
for (f in function_files) source(f, local = TRUE)

# Project configuration constants (from _targets.R)
STUDY_REGIONS <- c('Saint-Louis', 'Thiès', 'Fatick', 'Kaffrine')
ALT_STUDY_REGIONS <- c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')
MISSION_LABELS <- c('Mission 1 (July)', 'Mission 2 (September)', 'Mission 3 (October)')
DEFAULT_PALETTE <- "Degas"
FERTILIZER_COLORS <- c('control' = 'black', 'fertilized' = 'dark green')
DEFAULT_POINT_SIZE <- 5
DEFAULT_EMMEAN_POINT_SIZE <- 8
DEFAULT_FIGURE_DPI <- 2
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data from targets pipeline and filter for control plots only
data <- read_csv(here("data", "processed", "ose_data_processed.csv"), show_col_types = FALSE) |>
    filter(fertilizer_treatment == 'control') |>
    select(year, region, farmer, mission_number, ose_count, temperature) |>
    mutate(
        region = factor(region),
        farmer = factor(farmer),
        mission_number = factor(mission_number)
    )
```

# How is locust density related to temperature through the missions

We would like to model the influence of temperature on locust density. Let's plot the raw data by region to see how it falls. I also removed the fertilizer plots since this variable influences OSE counts.

```{r raw-data-viz}
#| echo: false
#| message: false
#| warning: false

raw_temp_plot <- data |>
    ggplot(aes(x = temperature, y = ose_count, color = region)) +
        geom_point() +
        geom_smooth(se = FALSE) +
        theme_pubr() +
        scale_color_met_d(name = 'Degas') +
        facet_wrap(~region, scales = 'free')

# Display the raw data plot
raw_temp_plot
```

The raw plots do not look very compelling, but let's construct a generalized additive model and see if there is a statistically significant trend.

```{r building-gam}
#| echo: false
#| message: false
#| warning: false

# Load temperature GAM from targets pipeline
mod <- readRDS(here::here("data", "model_objects", "locust_temperature_gam.rds"))
```

```{r}
#| echo: false
#| message: false
#| warning: false

# Extract smooth estimates for temperature by region
est <- gratia::smooth_estimates(mod, select = 'temperature', partial_match = TRUE)

# Create temperature smooth plot
temp_smooth_plot <- ggplot(est, aes(x = temperature, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - .se, ymax = .estimate + .se, fill = region), alpha = 0.3) +
  geom_line(aes(color = region)) +
  scale_fill_met_d(name = 'Degas') +
  scale_color_met_d(name = 'Degas') +
  labs(
    x = "Temperature (°C)",
    y = "Estimated OSE Count (Partial effects)",
    title = "Locust density × temperature"
  ) +
  theme_pubr()

temp_smooth_plot
```

There seems to be some positive realationship between temperature and locust denisty for Fatick, Saint Louis, and Thies. Kaffrine is a different relationship, however I wonder if this is partially still driven by the high densities in the first mission.

Here is the model summary for what it's worth:

```{r locust_temperature_model_summary}
#| echo: false
#| message: false
#| warning: false

model_summary_gt(mod, title = 'Locust x Temperature model summary', digits = 3) 
```



# New analysis (12/26/2025)

## Overview

I checked how temperature varied across missions and regions. Fatick heated up through the sampling period, but Kaffrine, Thies, and Saint Louis were more stable. Saint Louis was generally colder than the other regions.

I then ran a GAM with a flexible smooth for temperature by region, including mission number and farmer as random effects. The trends were pretty similar to the original model, with the exception of some odd patterns at cooler temps.

To simplify, I also ran a model with temperature as the main variable and region/mission as random effects. In this case, both region and mission number came up significant for explaining OSE counts, but temperature and farmer had little effect.

Overall, it seems like temperature at the time of the survey isn't a real driver for OSE counts once you control for region and mission—suggesting most of any pattern is noise or an artifact of timing/activity. I’m inclined to say temperature doesn’t need to be a focus in the main model, but happy to dig further if needed.


```{r}

data |>
    ggplot(aes(x=factor(mission_number),y=temperature)) +
    geom_jitter(width=0.2,height=0) +
    facet_wrap(~region) +
    xlab('mission number') +
    ylab('Temperature')


```


It looks like Fatick became hotter through the sampling period. Kaffrine, Thies, and Saint Louis seem more consistent between missions. Saint louis is definitely colder than the other three regions. So I would expect with the exception of Fatick, adding mission number to the model will not really add anything different.

Lets create the same model above with mission number included:

```{r new model}
#| eval: false


model_with_mission_number <- mgcv::gam(
            ose_count ~
                s(temperature, region, bs = 'fs', k = 10) +
                s(farmer, bs = 're') +
                s(mission_number, bs = 're'),
            data = data,
            family = mgcv::tw(),
            select = TRUE
        )

est <- gratia::smooth_estimates(model_with_mission_number,select = 'temperature',partial_match =TRUE)


ggplot(est, aes(x = temperature, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - .se, ymax = .estimate + .se,fill=region), alpha = 0.2) +
  geom_line(aes(color=region)) +
  labs(
    x = "Percent Ground Cover",
    y = "Estimated OSE Count (Partial effects)"
  ) +
  theme_minimal()
```



Seems like very similiar trends to the original model with the exception of more weird patterns in cooler temperatures. 


# Holding region and mission number as random effects

Lets try one more model where we do not try to look at the relationship between ose_count and region separately, but rather hold temperature as a random effect and look solely at temperature

```{r}
#| eval: false

temp_model <- mgcv::gam(
            ose_count ~
                s(temperature, k = 10) +
                s(region,bs='re') +
                s(farmer, bs = 're') +
                s(mission_number, bs = 're'),
            data = data,
            family = mgcv::tw(),
            select = TRUE
        )

summary(temp_model)

```



So this model has region and mission number being significant effects on ose_count which makes sense. However, temperature and farmer did not meaningfully impact ose_count. This makes me believe that we may be looking too deep into this relationship between ose_count and temperature and that we are fitting the model to pure noise. 

I am happy to work through this more, however I want to see what you think before going forward.