---
title: "2021 Locust × Ground Cover"
format: html
toc: true
---

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("locust_density_ground_cover")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data
data <- load_ose_data(cache_processed = TRUE)
```

We would like to see how density is influenced by ground cover. To do this I am building a generalized additive model (family: tweedie). Following this form:

The generalized additive model (GAM) fitted is:

$$
\begin{align*}
\text{ose\_count}_i = \\
  &\alpha + \\ 
  &+ f(\text{percent\_ground\_cover}_i,\, \text{region\_treat}_i) +  \\
  &+ b_{\text{farmer}_i} + \\
  &+ \varepsilon_i
\end{align*}
$$

where:

-   $\alpha$ is the intercept,
-   $f(\text{percent\_ground\_cover},\, \text{region\_treat})$ is a smooth function (tensor product, using factor-smooth interaction) over percent ground cover and region treatment,
-   $b_{\text{farmer}_i}$ is a random effect for each farmer,
-   $\varepsilon_i$ is the residual error.

```{r constructing-gam-model}
#| echo: false
#| message: false
#| warning: false

ground_cover_dat <- data |>
    dplyr::select(region, farmer, fertilizer_treatment, mission_number, ose_count, percent_ground_cover) |>
    mutate(
        region_treat = factor(paste0(region, '_', fertilizer_treatment)),
        farmer = factor(farmer)
    ) |>
    drop_na(ose_count, fertilizer_treatment)

# Use model caching for the GAM (can take several minutes to fit)
ground_cover_mod <- cache_model(
    model_name = "locust_ground_cover_gam",
    data = ground_cover_dat,
    fit_function = function(data, formula, nthreads, family) {
        build_gam(data, 
                 formula = formula, 
                 nthreads = nthreads, 
                 family = family)
    },
    # Simpler model for CI environments (faster fitting)
    fit_function_simple = function(data, formula, nthreads, family) {
        # Use simpler GAM with reduced complexity for CI
        build_gam(data,
                 formula = ose_count ~ te(percent_ground_cover, region_treat, bs = 'fs', k = 5) + s(farmer, bs = 're') +
                 mission_number,
                 nthreads = 1,
                 family = family)
    },
    formula = ose_count ~ te(percent_ground_cover, region_treat, bs = 'fs', k = 10) + s(farmer, bs = 're'),
    nthreads = 4,  # Reduced from 9 for better compatibility
    family = mgcv::tw()
)
```

The model diagnostics look pretty good! I tried to build a poisson model, but the data was overdispersed. Instead I went with a tweedie model which does a pretty good job at handling this type of data:

```{r gam model diagnostics}
#| echo: false
#| message: false
#| warning: false


appraise(ground_cover_mod)
```

However, I am not able to reproduce the smooths seen in the manuscript sent to me. This is likely due to data differences as I am not sure what was included in the initial modeling approach. **Roughly** we can see similiar trends in Kaffrine, Fatick, and Saint Louis. But Thies definitely has a different trend. Additionally, there seems to be different trends between control and fertilized treatments as comapred to the current plots in the manuscript. Not sure what to think here.

```{r plotting gam model results}
#| echo: false
#| message: false
#| warning: false

gam_plot <- plot_gam_smooths_gratia(ground_cover_mod)

gam_plot
```

## GAM model summary

```{r}
#| echo: false
#| message: false
#| warning: false


broom.mixed::tidy(ground_cover_mod) |>
  gt() |>
  fmt_number(
    columns = c(edf, ref.df, statistic, p.value),
    decimals = 2
  ) |>
  cols_label(
    term = "Term",
    edf = "Estimated DF",
    ref.df = "Reference DF",
    statistic = "F Statistic",
    p.value = "P-value"
  ) |>
  tab_header(
    title = "GAM model Summary: OSE Density x ground Cover"
  ) |>
  tab_options(
    table.font.size = "small",
    row.striping.background_color = "#F6F6EF"
  ) |>
  as_raw_html()


```


# New analysis (12/26/2025)

I built a GAM with a Tweedie family and log link to model `ose_count`. The model includes a smooth for percent ground cover and random effects for region, fertilizer treatment, mission number, and farmer. 

The relationship between ground cover and OSE count is non-linear, peaking at around 50% ground cover—OSE numbers are lower at both low and high cover. Region, fertilizer, mission, and farmer all accounted for substantial variation. 

Overall, the model explains about 52% of the deviance in the data.  



```{r running_new_model}


data <- data |>
    select(year,region,farmer,farmer_gender,fertilizer_treatment,mission_number,ose_count,percent_ground_cover) |>
    mutate(fertilizer_treatment = factor(fertilizer_treatment),
           region = factor(region),
           farmer = factor(farmer),
           mission_number = factor(mission_number),
           region_treat = factor(paste0(region, '_', fertilizer_treatment))
           )

data |> 
  glimpse()


mod <- bam(
  ose_count ~ 
    s(percent_ground_cover) + 
    s(region, bs="re") +
    s(fertilizer_treatment, bs="re") +
    s(mission_number, bs="re") +
    s(farmer, bs="re"),
  data = data,
  family = tw()
)

gratia::appraise(mod)

```



Here is the final modeled results for the relationship between ose_count and ground cover


```{r plotting_estimates}



est <- smooth_estimates(mod, select = "percent_ground_cover",partial_match=TRUE)

est$adj_est <- est$.estimate + coef(mod)[1]


ggplot(est, aes(x = percent_ground_cover, y = adj_est)) +
  geom_ribbon(aes(ymin = adj_est - .se, ymax = adj_est + .se), alpha = 0.2) +
  geom_line() +
  labs(
    x = "Percent Ground Cover",
    y = "Estimated OSE Count"
  ) +
  theme_minimal()



```


# ground cover by mission

in general, ground cover increased through the year. It does not really seem theres a big difference between regions.

```{r}

data |>
  ggplot(aes(x=mission_number,y=percent_ground_cover)) +
  geom_boxplot(outlier.size=-1) +
  geom_jitter(width=0.2,height=0) +
  facet_wrap(~region) +
  ggtitle('ground cover by region') +
  xlab('mission number') +
  ylab('ground cover (%)')


```

