---
title: "2021 Yield × Treatment × Region Results"
subtitle: "Effects of fertilization on millet yield by region"
format: html
toc: true
---

::: callout-note
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

# Goal

This analysis examines how fertilization influences millet yield across different regions. We expect to see regional differences in fertilizer response.

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Core packages for visualization
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(ggpubr)
  library(glmmTMB)
  library(emmeans)
  library(DHARMa)
  library(broom.mixed)
  library(gt)
  library(webshot2)
})

# Plot defaults
theme_set(theme_pubr())
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.retina = 2
)

# Source project functions
function_files <- list.files(here("R", "functions"), pattern = "\\.R$", full.names = TRUE)
for (f in function_files) source(f, local = TRUE)

# Project configuration constants (from _targets.R)
STUDY_REGIONS <- c('Saint-Louis', 'Thiès', 'Fatick', 'Kaffrine')
ALT_STUDY_REGIONS <- c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')
MISSION_LABELS <- c('Mission 1 (July)', 'Mission 2 (September)', 'Mission 3 (October)')
DEFAULT_PALETTE <- "Degas"
FERTILIZER_COLORS <- c('control' = 'black', 'fertilized' = 'dark green')
DEFAULT_POINT_SIZE <- 5
DEFAULT_EMMEAN_POINT_SIZE <- 8
DEFAULT_FIGURE_DPI <- 2

# Helper function to save gt tables as PNG and CSV
save_gt_png <- function(gt_obj, filename, subfolder = "yield_treatment_region") {
  # Create output directories
  png_dir <- here::here("outputs", "tables", subfolder, "png")
  csv_dir <- here::here("outputs", "tables", subfolder, "csv")
  dir.create(png_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Save PNG
  png_path <- file.path(png_dir, filename)
  gtsave(gt_obj, png_path, vwidth = 3000)
  
  # Save CSV - extract the underlying data from the gt object
  csv_filename <- sub("\\.png$", ".csv", filename)
  csv_path <- file.path(csv_dir, csv_filename)
  
  # Extract data from gt object
  gt_data <- gt_obj$`_data`
  write_csv(gt_data, csv_path)
  
  return(invisible(list(png = png_path, csv = csv_path)))
}
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data from targets pipeline and filter for yield analysis
data <- read_csv(here("data", "processed", "ose_data_processed.csv"), show_col_types = FALSE) |>
    select(year, region, farmer, fertilizer_treatment, rendement_en_kg_ha) |>
    drop_na(rendement_en_kg_ha) |>
    distinct()
```

# Raw data visualization

It is fairly obvious that fertilization influences millet yield. It is interesting to see the regional differences. It appears that Thies has a much strong response to fertilization than farms at Fatick or Kaffrine. Or at least the variation is dramatically less.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
avg_yield <- data |>
    group_by(region,fertilizer_treatment) |>
    summarize(mean_yield = mean(rendement_en_kg_ha)) |>
    drop_na() 

data |>
    drop_na() |>
    ggplot(aes(x=region,y=rendement_en_kg_ha,color=fertilizer_treatment)) +
        geom_jitter(position=position_jitterdodge()) +
        geom_jitter(data = avg_yield,aes(y=mean_yield),size=8,position=position_jitterdodge(jitter.width=0,jitter.height=0)) +
        scale_color_manual(values = c('black','dark green')) +
        theme_pubr() +
        xlab('') +
        ylab('yield')         

```

# Modeling

This model formula will be similar to locust density by treatment x region. I will hold farmer as a random effect. So the model formula looks like this:

```{r model formula}

formula <- yield ~
    fertilizer_treatment * region + # interactive fixed effect
    (1|farmer) # random effect

```

```{r model-building}
#| echo: false
#| message: false
#| warning: false

# Load yield model from targets pipeline
yield_mod <- readRDS(here::here("data", "model_objects", "yield_treatment_region.rds"))

# Define emmeans function locally (could be moved to statistical_models.R)
get_emmeans_and_pairs <- function(model) {
  emmeans_obj <- emmeans::emmeans(model, ~ fertilizer_treatment * region, type = "response")
  emmeans_tbl <- as_tibble(emmeans_obj)

  pairs_tbl <- as_tibble(pairs(emmeans_obj, adjust = "tukey"))

  list(
    emmeans = emmeans_tbl,
    pairwise = pairs_tbl
  )
}
```

As can be seen in the modeled results below, fertilization significantly improved yield across the board. For whatever reason, Thies had lower yield unfertilized but also saw the biggest incase in yield when fertilized.

```{r mod emmeans}
#| echo: false
#| message: false
#| warning: false

emms <- get_emmeans_and_pairs(yield_mod)

emms$emmeans |>
    ggplot(aes(x=region,y=response,color=fertilizer_treatment)) +
        geom_point(position = position_dodge(width=0.2),size=10) +
        geom_errorbar(aes(ymin=asymp.LCL, ymax=asymp.UCL),width=0,position = position_dodge(width=0.2)) +
        scale_color_manual(values=c('black','dark green')) +
        theme_pubr() +
        xlab('') +
        ylab('yield (estimated marginal means)')


```

Here are the tabulated summaries for the model:

```{r model summary}
#| echo: false
#| message: false
#| warning: false

tidy_prepped <- broom.mixed::tidy(yield_mod) %>%
  # Make term names human-readable; handle interactions by replacing ":" with " × "
  mutate(
    term = term %>%
      # common renames from your output
      str_replace_all("^\\(Intercept\\)$", "Intercept") %>%
      str_replace_all("fertilizer_treatmentfertilized", "Fertilized") %>%
      str_replace_all("regionKaffrine", "Kaffrine") %>%
      str_replace_all("regionThies", "Thies"),
    # keep original p.value column but create a formatted label with significance stars
    p_value = p.value,
    p_label = case_when(
      is.na(p_value) ~ NA_character_,
      p_value < 0.001 ~ "<0.001",
      TRUE ~ formatC(p_value, digits = 3, format = "f")
    ),
    sig = case_when(
      is.na(p_value) ~ "",
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p_label = if_else(is.na(p_label), NA_character_, paste0(p_label, sig))
  ) %>%
  # Keep ordering as-is; you can reorder rows here if desired
  select(effect, component, term, estimate, std.error, statistic, p_value, p_label)

# ---- Build the gt table ----
gt_table <- tidy_prepped %>%
  # start GT table
  gt(rowname_col = "term") %>%
  # Add row groups for fixed vs random parts
  tab_row_group(
    group = "Fixed effects",
    rows = tidy_prepped$effect == "fixed"
  ) %>%
  tab_row_group(
    group = "Random parameters",
    rows = tidy_prepped$effect == "ran_pars"
  ) %>%
  # Column labels for presentation
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. error",
    statistic = "z",
    p_label = "p-value",
    component = "Component"
  ) %>%
  # Format numeric columns
  fmt_number(
    columns = vars(estimate, std.error, statistic,p_value),
    decimals = 3
  ) %>%
  # Replace NA with a dash
  fmt_missing(
    columns = everything(),
    missing_text = "-"
  ) %>%
  # Move the p-value column to the rightmost position
  cols_move_to_end(columns = vars(p_label)) %>%
  # Styling: bold the row group labels
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  # Header and source note suitable for a Quarto site
  tab_header(
    title = md("Model estimates for yield ~ fertilizer x region model"),
    subtitle = md("Fixed effects and random parameter SDs")
  ) %>%
  # Accessibility / alignment tweaks
  cols_align(
    align = "center",
    columns = vars(estimate, std.error, statistic, p_label)
  )

# Print or return the table
save_gt_png(gt_table, "yield_fertilizer_region_model_summary.png")
gt_table

```

And of course the pairwise comparisons for `fertilizer x region`

```{r}

# Prepare and format p-value with significance stars
emms_pw_prepped <- emms$pairwise %>%
  mutate(
    p_label = case_when(
      is.na(p.value)   ~ "-",
      p.value == 0     ~ "<0.001***",
      p.value < 0.001  ~ "<0.001***",
      p.value < 0.01   ~ paste0(formatC(p.value, digits = 3, format = "e"), "**"),
      p.value < 0.05   ~ paste0(formatC(p.value, digits = 3, format = "e"), "*"),
      TRUE             ~ formatC(p.value, digits = 3, format = "e")
    )
  )

# Make column names readable
col_labels <- c(
  contrast   = "Contrast",
  ratio      = "Ratio",
  SE         = "Std. Error",
  df         = "DF",
  null       = "Null",
  z.ratio    = "z Ratio",
  p_label    = "p-value"
)



emms_pairwise_gt <- emms_pw_prepped %>%
  select(contrast, ratio, SE, df, null, z.ratio, p_label) %>%
  gt(rowname_col = "contrast") %>%
  cols_label(!!!col_labels) %>%
  fmt_number(
    columns = c("ratio", "SE", "z.ratio"),
    decimals = 3
  ) %>%
  fmt_missing(columns = everything(), missing_text = "-") %>%
  tab_header(
    title = md("Estimated Marginal Means Pairwise Contrasts"),
    subtitle = md("Table of pairwise ratios, z statistics, and p-values")
  ) %>%
  cols_align(
    align = "center",
    columns = c("ratio", "SE", "z.ratio", "null", "df", "p_label")
  ) %>%
  opt_row_striping()

save_gt_png(emms_pairwise_gt, "yield_fertilizer_region_pairwise_comparisons.png")
emms_pairwise_gt

# Print or return the table (Quarto HTML will render this table)
emms_pairwise_gt


```