---
title: "2021 Yield × Locust Results"
subtitle: "Impact of locust density on millet yield"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

# Goal

We would like to see if locust density influences end-of-season millet yield. Locust density was collected three times through the season and yield is collected at the end of season.

Like the first paper, we are running into confounding variable issues specifically with fertilization. To overcome this, I will focus on the control plots only.

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Core packages for visualization
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(ggpubr)
  library(glmmTMB)
  library(emmeans)
  library(DHARMa)
  library(broom.mixed)
  library(gt)
  library(webshot2)
})

# Plot defaults
theme_set(theme_pubr())
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.retina = 2
)

# Source project functions
function_files <- list.files(here("R", "functions"), pattern = "\\.R$", full.names = TRUE)
for (f in function_files) source(f, local = TRUE)

# Project configuration constants (from _targets.R)
STUDY_REGIONS <- c('Saint-Louis', 'Thiès', 'Fatick', 'Kaffrine')
ALT_STUDY_REGIONS <- c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')
MISSION_LABELS <- c('Mission 1 (July)', 'Mission 2 (September)', 'Mission 3 (October)')
DEFAULT_PALETTE <- "Degas"
FERTILIZER_COLORS <- c('control' = 'black', 'fertilized' = 'dark green')
DEFAULT_POINT_SIZE <- 5
DEFAULT_EMMEAN_POINT_SIZE <- 8
DEFAULT_FIGURE_DPI <- 2

# Helper function to save gt tables as PNG and CSV
save_gt_png <- function(gt_obj, filename, subfolder = "yield_locust") {
  # Create output directories
  png_dir <- here::here("outputs", "tables", subfolder, "png")
  csv_dir <- here::here("outputs", "tables", subfolder, "csv")
  dir.create(png_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Save PNG
  png_path <- file.path(png_dir, filename)
  gtsave(gt_obj, png_path, vwidth = 3000)
  
  # Save CSV - extract the underlying data from the gt object
  csv_filename <- sub("\\.png$", ".csv", filename)
  csv_path <- file.path(csv_dir, csv_filename)
  
  # Extract data from gt object
  gt_data <- gt_obj$`_data`
  write_csv(gt_data, csv_path)
  
  return(invisible(list(png = png_path, csv = csv_path)))
}
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data from targets pipeline and filter for yield analysis
data <- read_csv(here("data", "processed", "ose_data_processed.csv"), show_col_types = FALSE) |>
    filter(fertilizer_treatment == 'control') |>
    select(year, region, farmer, mission_number, fertilizer_treatment, ose_count, rendement_en_kg_ha) |>
    drop_na(rendement_en_kg_ha)
```

## Raw data visualization


```{r raw data viz}
#| echo: false
#| message: false
#| warning: false


lm_plot <- data |>
    group_by(region,farmer) |>
    mutate(mean_ose = mean(ose_count)) |>
    ggplot(aes(x=ose_count,y=rendement_en_kg_ha,color=region)) +
        geom_point(pch=21) +
        geom_smooth(method='lm',se=FALSE,size=1.25) +
        theme_pubr() +
        xlab('OSE count') +
        ylab('yield') +
        MetBrewer::scale_color_met_d(name = 'Degas') +
        facet_wrap(~region,ncol=2,scale='free')

lm_plot
```

Doesnt look too strong, but lets model. I will hold the following variables as random effects: *farmer number*, *region*, and *mission_number*. Holding mission_number as a random effect can help with puesodreplication. 


```{r model-building}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load yield-locust model from targets pipeline
mod <- readRDS(here::here("data", "model_objects", "yield_locust_density.rds"))
```

There isnt a strong relationship between locust density and yield. I am happy to dive further into this if you'd like!


```{r model viz}
#| echo: false
#| message: false
#| warning: false
#| 


tidy_prepped <- broom.mixed::tidy(mod) %>%
  # Make labels friendlier
  mutate(
    Effect = case_when(
      effect == "fixed" ~ "Fixed effects",
      effect == "ran_pars" ~ "Random parameters",
      TRUE ~ effect
    ),
    Component = if_else(is.na(component) | component == "NA", "Overall", as.character(component)),
    Group = if_else(is.na(group) | group == "NA", "-", as.character(group)),
    # Clean up term names, turn sd__(...) into "SD — <what>" for readability
    Term = case_when(
      term == "(Intercept)" ~ "Intercept",
      str_detect(term, "^sd__") ~ str_replace(term, "^sd__\\(?([^\\)]*)\\)?", "SD — \\1"),
      TRUE ~ as.character(term)
    ),
    # Prepare a pretty p-value column for display
    p_display = case_when(
      is.na(p.value) ~ NA_character_,
      p.value == 0 ~ "<0.001",
      p.value < 0.001 ~ paste0("<", formatC(0.001, format = "f", digits = 3)),
      TRUE ~ formatC(p.value, format = "f", digits = 3)
    )
  ) %>%
  # Choose ordering for the table: fixed effects first, then random parameters
  mutate(effect_order = case_when(Effect == "Fixed effects" ~ 1L, Effect == "Random parameters" ~ 2L, TRUE ~ 3L)) %>%
  arrange(effect_order, Effect, Component, Group)

# Build gt table
tbl_yield_locust <- tidy_prepped %>%
  select(Effect, Component, Group, Term, Estimate = estimate, `Std. Error` = std.error, Statistic = statistic, `p value` = p_display) %>%
  gt(rowname_col = "Term") %>%
  # Create row groups by Effect (Fixed / Random)
  tab_row_group(
    group = "Fixed effects",
    rows = Effect == "Fixed effects"
  ) %>%
  tab_row_group(
    group = "Random parameters",
    rows = Effect == "Random parameters"
  ) %>%
  # Column labels and title
  cols_label(
    Component = "Component",
    Group = "Group",
    `Std. Error` = "Std. Error",
    Statistic = "Statistic",
    `p value` = "p value"
  ) %>%
  tab_header(
    title = md("Model coefficients"),
    subtitle = md("Fixed effects and estimated random-parameter standard deviations")
  ) %>%
  # Format numeric columns:
  # - For Estimate: show normal decimals for medium/large values, scientific for very small
  fmt_number(
    columns = vars(Estimate),
    decimals = 3,
    rows = !is.na(Estimate) & abs(Estimate) >= 0.001
  ) %>%
  fmt_scientific(
    columns = vars(Estimate),
    rows = !is.na(Estimate) & abs(Estimate) < 0.001,
    decimals = 2
  ) %>%
  fmt_number(
    columns = vars(`Std. Error`, Statistic),
    decimals = 3
  ) %>%
  # Replace missing numeric entries with a dash for clarity
  fmt_missing(
    columns = everything(),
    missing_text = "-"
  ) %>%
  # A little visual polish
  opt_align_table_header("left")

save_gt_png(tbl_yield_locust, "yield_locust_density_model_summary.png")
tbl_yield_locust
```