---
title: "2021 Yield x Temperature"
subtitle: "figures and tables"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

# Goal

How is yield millet influenced by temperature during the growing season? This may be overkill, but I thought I would show this as well since the data has already been collected.

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Core packages for visualization
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(ggpubr)
  library(glmmTMB)
  library(mgcv)
})

# Plot defaults
theme_set(theme_pubr())
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.retina = 2
)

# Source project functions
function_files <- list.files(here("R", "functions"), pattern = "\\.R$", full.names = TRUE)
for (f in function_files) source(f, local = TRUE)

# Project configuration constants (from _targets.R)
STUDY_REGIONS <- c('Saint-Louis', 'Thiès', 'Fatick', 'Kaffrine')
ALT_STUDY_REGIONS <- c('Saint Louis', 'Thies', 'Fatick', 'Kaffrine')
MISSION_LABELS <- c('Mission 1 (July)', 'Mission 2 (September)', 'Mission 3 (October)')
DEFAULT_PALETTE <- "Degas"
FERTILIZER_COLORS <- c('control' = 'black', 'fertilized' = 'dark green')
DEFAULT_POINT_SIZE <- 5
DEFAULT_EMMEAN_POINT_SIZE <- 8
DEFAULT_FIGURE_DPI <- 2
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data from targets pipeline
data <- read_csv(here("data", "processed", "ose_data_processed.csv"), show_col_types = FALSE) |>
    select(year, region, farmer, temperature, fertilizer_treatment, rendement_en_kg_ha) |>
    drop_na(rendement_en_kg_ha, fertilizer_treatment) |>
    distinct()
```

# Raw data visualization

So I will calculate the mean and sd temperature and plot against yield to see if there are any interesting trends.

```{r}

data |>
    group_by(region,farmer) |>
    summarize(mean_temp = mean(temperature),
        sd_temp = sd(temperature),
        rendement_en_kg_ha = first(rendement_en_kg_ha)) |>
    pivot_longer(!c(farmer,region,rendement_en_kg_ha)) |>
    ggplot(aes(x=value,y=rendement_en_kg_ha)) +
        geom_point() +
        geom_smooth(method='gam',se=FALSE) +
        facet_grid(region~name,scales='free') +
        theme_pubr()


```

Not sure what to think about the potential non-linear relationship with yield x temperature. It might not be significant anyway. Happy to go further here if y'all are interested but for now I will leave it.

