---
title: "2021 Yield × Locust Results"
subtitle: "Impact of locust density on millet yield"
format: html
toc: true
---

::: {.callout-note}
## Exploratory Work

This page contains exploratory analysis conducted during the development of the manuscript. These analyses informed our statistical approach and data management decisions. Not all content here appears in the final manuscript. For the final methods and results, see the [Methods and Results](manuscript_methods_results.qmd) page.
:::

# Goal

We would like to see if locust density influences end-of-season millet yield. Locust density was collected three times through the season and yield is collected at the end of season.

Like the first paper, we are running into confounding variable issues specifically with fertilization. To overcome this, I will focus on the control plots only.

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("yield_locust")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data and filter for yield analysis
data <- load_ose_data(cache_processed = TRUE) |>
    filter(fertilizer_treatment == 'control') |>
    select(year, region, farmer, mission_number, fertilizer_treatment, ose_count, rendement_en_kg_ha) |>
    drop_na(rendement_en_kg_ha)
```

## Raw data visualization


```{r raw data viz}
#| echo: false
#| message: false
#| warning: false


lm_plot <- data |>
    group_by(region,farmer) |>
    mutate(mean_ose = mean(ose_count)) |>
    ggplot(aes(x=ose_count,y=rendement_en_kg_ha,color=region)) +
        geom_point(pch=21) +
        geom_smooth(method='lm',se=FALSE,size=1.25) +
        theme_pubr() +
        xlab('OSE count') +
        ylab('yield') +
        MetBrewer::scale_color_met_d(name = 'Degas') +
        facet_wrap(~region,ncol=2,scale='free')

lm_plot
```

Doesnt look too strong, but lets model. I will hold the following variables as random effects: *farmer number*, *region*, and *mission_number*. Holding mission_number as a random effect can help with puesodreplication. 


```{r model-building}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Use model caching for the yield-locust model
mod <- cache_model(
    model_name = "yield_locust_density",
    data = data,
    fit_function = function(data) {
        glmmTMB(
            rendement_en_kg_ha ~ 
                ose_count +
                (1|region) +
                (1|farmer) +
                (1|mission_number),
            data = data,
            family = poisson()
        )
    }
)
```

There isnt a strong relationship between locust density and yield. I am happy to dive further into this if you'd like!


```{r model viz}
#| echo: false
#| message: false
#| warning: false
#| 


tidy_prepped <- broom.mixed::tidy(mod) %>%
  # Make labels friendlier
  mutate(
    Effect = case_when(
      effect == "fixed" ~ "Fixed effects",
      effect == "ran_pars" ~ "Random parameters",
      TRUE ~ effect
    ),
    Component = if_else(is.na(component) | component == "NA", "Overall", as.character(component)),
    Group = if_else(is.na(group) | group == "NA", "-", as.character(group)),
    # Clean up term names, turn sd__(...) into "SD — <what>" for readability
    Term = case_when(
      term == "(Intercept)" ~ "Intercept",
      str_detect(term, "^sd__") ~ str_replace(term, "^sd__\\(?([^\\)]*)\\)?", "SD — \\1"),
      TRUE ~ as.character(term)
    ),
    # Prepare a pretty p-value column for display
    p_display = case_when(
      is.na(p.value) ~ NA_character_,
      p.value == 0 ~ "<0.001",
      p.value < 0.001 ~ paste0("<", formatC(0.001, format = "f", digits = 3)),
      TRUE ~ formatC(p.value, format = "f", digits = 3)
    )
  ) %>%
  # Choose ordering for the table: fixed effects first, then random parameters
  mutate(effect_order = case_when(Effect == "Fixed effects" ~ 1L, Effect == "Random parameters" ~ 2L, TRUE ~ 3L)) %>%
  arrange(effect_order, Effect, Component, Group)

# Build gt table
gt_tbl <- tidy_prepped %>%
  select(Effect, Component, Group, Term, Estimate = estimate, `Std. Error` = std.error, Statistic = statistic, `p value` = p_display) %>%
  gt(rowname_col = "Term") %>%
  # Create row groups by Effect (Fixed / Random)
  tab_row_group(
    group = "Fixed effects",
    rows = Effect == "Fixed effects"
  ) %>%
  tab_row_group(
    group = "Random parameters",
    rows = Effect == "Random parameters"
  ) %>%
  # Column labels and title
  cols_label(
    Component = "Component",
    Group = "Group",
    `Std. Error` = "Std. Error",
    Statistic = "Statistic",
    `p value` = "p value"
  ) %>%
  tab_header(
    title = md("Model coefficients"),
    subtitle = md("Fixed effects and estimated random-parameter standard deviations")
  ) %>%
  # Format numeric columns:
  # - For Estimate: show normal decimals for medium/large values, scientific for very small
  fmt_number(
    columns = vars(Estimate),
    decimals = 3,
    rows = !is.na(Estimate) & abs(Estimate) >= 0.001
  ) %>%
  fmt_scientific(
    columns = vars(Estimate),
    rows = !is.na(Estimate) & abs(Estimate) < 0.001,
    decimals = 2
  ) %>%
  fmt_number(
    columns = vars(`Std. Error`, Statistic),
    decimals = 3
  ) %>%
  # Replace missing numeric entries with a dash for clarity
  fmt_missing(
    columns = everything(),
    missing_text = "-"
  ) %>%
  # A little visual polish
  opt_align_table_header("left")


gt_tbl
```