---
title: "2021 Yield x Temperature"
subtitle: "figures and tables"
format: html
toc: true
---

# Goal

How is yield millet influenced by temperature during the growing season? This may be overkill, but I thought I would show this as well since the data has already been collected.

```{r setup}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

# Load common setup (packages, functions, configuration)
source(here::here("R", "_common.R"))

# Set the current analysis for organized output
set_current_analysis("yield_environment")
```

```{r data-loading}
#| echo: false
#| message: false
#| warning: false

# Load processed data
data <- load_ose_data(cache_processed = TRUE) |>
    select(year, region, farmer, temperature, fertilizer_treatment, rendement_en_kg_ha) |>
    drop_na(rendement_en_kg_ha, fertilizer_treatment) |>
    distinct()
```

# Raw data visualization

So I will calculate the mean and sd temperature and plot against yield to see if there are any interesting trends.

```{r}

data |>
    group_by(region,farmer) |>
    summarize(mean_temp = mean(temperature),
        sd_temp = sd(temperature),
        rendement_en_kg_ha = first(rendement_en_kg_ha)) |>
    pivot_longer(!c(farmer,region,rendement_en_kg_ha)) |>
    ggplot(aes(x=value,y=rendement_en_kg_ha)) +
        geom_point() +
        geom_smooth(method='gam',se=FALSE) +
        facet_grid(region~name,scales='free') +
        theme_pubr()


```

Not sure what to think about the potential non-linear relationship with yield x temperature. It might not be significant anyway. Happy to go further here if y'all are interested but for now I will leave it.

