---
title: "2021 Yield x treatment X region results"
subtitle: "figures and tables"
format: html
toc: true
---

# Goal

This should be a straight forward investigation as we are simply looking at how fertilization inlfluences millet yield. We should look at it by region to see any regional differences.

```{r package loading}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false


library(tidyverse)
library(rnaturalearth)
library(rnaturalearthhires)
library(janitor)
library(MetBrewer)
library(here)
library(patchwork)
library(ggpubr)
library(gt)
library(glmmTMB)
library(mgcv)
library(gratia)
library(emmeans)

here::i_am('README.md')

# setting project root
# sourcing functions
source(('R/functions/data_preprocessing.R'), local = knitr::knit_global())
source(('R/functions/table_variable_summary.R'), local = knitr::knit_global())
source(('R/functions/locust_density_temperature.R'), local = knitr::knit_global())
source(('R/functions/locust_density_modeling.R'), local = knitr::knit_global())

```

```{r data loading and processing}
#| echo: false
#| message: false
#| warning: false

data = process_senegal_data('data/raw/Toure_OSE2021data_v03.xlsx') |>
    select(year,region,farmer,fertilizer_treatment,rendement_en_kg_ha) |>
    drop_na(rendement_en_kg_ha) |>
    distinct()

```

# Raw data visualization

It is fairly obvious that fertilization influences millet yield. It is interesting to see the regional differences. It appears that Thies has a much strong response to fertilization than farms at Fatick or Kaffrine. Or at least the variation is dramatically less.

```{r}
#| echo: false
#| message: false
#| warning: false
#| 
avg_yield <- data |>
    group_by(region,fertilizer_treatment) |>
    summarize(mean_yield = mean(rendement_en_kg_ha)) |>
    drop_na() 

data |>
    drop_na() |>
    ggplot(aes(x=region,y=rendement_en_kg_ha,color=fertilizer_treatment)) +
        geom_jitter(position=position_jitterdodge()) +
        geom_jitter(data = avg_yield,aes(y=mean_yield),size=8,position=position_jitterdodge(jitter.width=0,jitter.height=0)) +
        scale_color_manual(values = c('black','dark green')) +
        theme_pubr() +
        xlab('') +
        ylab('yield')         

```

# Modeling

This model formula will be similar to locust density by treatment x region. I will hold farmer as a random effect. So the model formula looks like this:

```{r model formula}

formula <- yield ~
    fertilizer_treatment * region + # interactive fixed effect
    (1|farmer) # random effect

```

```{r model building}
#| echo: false
#| message: false
#| warning: false

yield_mod <- glmmTMB(
    rendement_en_kg_ha ~
        fertilizer_treatment * region +
        (1|farmer),
    data = data,
    family=tweedie()
)

get_emmeans_and_pairs <- function(glmm_mod) {
  emmeans_obj <- emmeans(glmm_mod, ~ fertilizer_treatment * region, type = "response")
  
  emmeans_tbl <- as_tibble(emmeans_obj) 
  
  pairs_tbl <- as_tibble(pairs(emmeans_obj, adjust = "tukey"))
  
  list(
    emmeans = emmeans_tbl,
    pairwise = pairs_tbl
  )
}


```

As can be seen in the modeled results below, fertilization significantly improved yield across the board. For whatever reason, Thies had lower yield unfertilized but also saw the biggest incase in yield when fertilized.

```{r mod emmeans}
#| echo: false
#| message: false
#| warning: false

emms <- get_emmeans_and_pairs(yield_mod)

emms$emmeans |>
    ggplot(aes(x=region,y=response,color=fertilizer_treatment)) +
        geom_point(position = position_dodge(width=0.2),size=10) +
        geom_errorbar(aes(ymin=asymp.LCL, ymax=asymp.UCL),width=0,position = position_dodge(width=0.2)) +
        scale_color_manual(values=c('black','dark green')) +
        theme_pubr() +
        xlab('') +
        ylab('yield (estimated marginal means)')


```

Here are the tabulated summaries for the model:

```{r model summary}
#| echo: false
#| message: false
#| warning: false

tidy_prepped <- broom.mixed::tidy(yield_mod) %>%
  # Make term names human-readable; handle interactions by replacing ":" with " × "
  mutate(
    term = term %>%
      # common renames from your output
      str_replace_all("^\\(Intercept\\)$", "Intercept") %>%
      str_replace_all("fertilizer_treatmentfertilized", "Fertilized") %>%
      str_replace_all("regionKaffrine", "Kaffrine") %>%
      str_replace_all("regionThies", "Thies"),
    # keep original p.value column but create a formatted label with significance stars
    p_value = p.value,
    p_label = case_when(
      is.na(p_value) ~ NA_character_,
      p_value < 0.001 ~ "<0.001",
      TRUE ~ formatC(p_value, digits = 3, format = "f")
    ),
    sig = case_when(
      is.na(p_value) ~ "",
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p_label = if_else(is.na(p_label), NA_character_, paste0(p_label, sig))
  ) %>%
  # Keep ordering as-is; you can reorder rows here if desired
  select(effect, component, term, estimate, std.error, statistic, p_value, p_label)

# ---- Build the gt table ----
gt_table <- tidy_prepped %>%
  # start GT table
  gt(rowname_col = "term") %>%
  # Add row groups for fixed vs random parts
  tab_row_group(
    group = "Fixed effects",
    rows = tidy_prepped$effect == "fixed"
  ) %>%
  tab_row_group(
    group = "Random parameters",
    rows = tidy_prepped$effect == "ran_pars"
  ) %>%
  # Column labels for presentation
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. error",
    statistic = "z",
    p_label = "p-value",
    component = "Component"
  ) %>%
  # Format numeric columns
  fmt_number(
    columns = vars(estimate, std.error, statistic,p_value),
    decimals = 3
  ) %>%
  # Replace NA with a dash
  fmt_missing(
    columns = everything(),
    missing_text = "-"
  ) %>%
  # Move the p-value column to the rightmost position
  cols_move_to_end(columns = vars(p_label)) %>%
  # Styling: bold the row group labels
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  # Header and source note suitable for a Quarto site
  tab_header(
    title = md("Model estimates for yield ~ fertilizer x region model"),
    subtitle = md("Fixed effects and random parameter SDs")
  ) %>%
  tab_source_note(
    source_note = md("Significance: *** p < 0.001, ** p < 0.01, * p < 0.05")
  ) %>%
  # Accessibility / alignment tweaks
  cols_align(
    align = "center",
    columns = vars(estimate, std.error, statistic, p_label)
  )

# Print or return the table
gt_table

```

And of course the pairwise comparisons for `fertilizer x region`

```{r}

# Prepare and format p-value with significance stars
emms_pw_prepped <- emms$pairwise %>%
  mutate(
    p_label = case_when(
      is.na(p.value)   ~ "-",
      p.value == 0     ~ "<0.001***",
      p.value < 0.001  ~ "<0.001***",
      p.value < 0.01   ~ paste0(formatC(p.value, digits = 3, format = "e"), "**"),
      p.value < 0.05   ~ paste0(formatC(p.value, digits = 3, format = "e"), "*"),
      TRUE             ~ formatC(p.value, digits = 3, format = "e")
    )
  )

# Make column names readable
col_labels <- c(
  contrast   = "Contrast",
  ratio      = "Ratio",
  SE         = "Std. Error",
  df         = "DF",
  null       = "Null",
  z.ratio    = "z Ratio",
  p_label    = "p-value"
)



emms_pairwise_gt <- emms_pw_prepped %>%
  select(contrast, ratio, SE, df, null, z.ratio, p_label) %>%
  gt(rowname_col = "contrast") %>%
  cols_label(!!!col_labels) %>%
  fmt_number(
    columns = c("ratio", "SE", "z.ratio"),
    decimals = 3
  ) %>%
  fmt_missing(columns = everything(), missing_text = "-") %>%
  tab_header(
    title = md("Estimated Marginal Means Pairwise Contrasts"),
    subtitle = md("Table of pairwise ratios, z statistics, and p-values")
  ) %>%
  tab_source_note(
    source_note = md("Significance: *** p < 0.001, ** p < 0.01, * p < 0.05")
  ) %>%
  cols_align(
    align = "center",
    columns = c("ratio", "SE", "z.ratio", "null", "df", "p_label")
  ) %>%
  opt_row_striping()

# Print or return the table (Quarto HTML will render this table)
emms_pairwise_gt


```